---
title: "Data Cleaning"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(atlantiscas)
library(magrittr)
```

The fishing data is a combination of VTR (1996-2007) and DMIS (2008-2021) data.

```{r sampleData, echo = F}
# read sample data and display it in a table
sampleData <- readRDS(here::here("data-raw/sampleFishingData.rds")) %>% 
  dplyr::select(Year,PORT,STATE,GEARCODE,GEARCAT, NESPP3,SPPNM)
DT::datatable(sampleData)
```


Both have different lookup tables and formatting for species names (SPPNM), port names (PORT), gear categories (GEARCAT)

## Ports

* Port lok up tables lack latitude and longitude
  * Use port, county, state to geocode lat and lon using open street maps
* Ports have duplicate entries with spelling typos
  * eg. harwichport, harwich port. These were hand fixed
  
  
All PORTID within the Atlantis domain were identified. Other PORTIDs were grouped into either Canadian ports (north of the atlantis domain) or southern ports (south of atlantis domain)

PORTID was used to match PORTNM,COUNTY,STATE from the CFDBS.PORT table (VTR.VLPORTSYN contains too many typos). Port names were then used to find lat and lon values using `tidygeocoder` usig open street maps and online service `geocode.maps.co`. Many port names and/or counties had to be edited due to spelling errors, incorrect information, inconsistent formatting, or missing information. For example:

  * SALISBURY COVE, ME -> SALSBURY COVE, ME (spelling)
  * (TOWN OF) -> (inconsistent formatting -removed text "(TOWN OF)" from port name)
  * PERKINS COVE, CUMBERLAND, ME -> OGUNQUIT, YORK, ME (incorrect)
  * OTHER FAIRFIELD, FAIRFIELD, CT -> FAIRFIELD, FAIRFIELD, CT (missing port info)
  * COMBS CREEK, ST. MARYS, MD -> NA, ST.MARY'S, MD (not found on maps)
  * DUKES, DUKES, MA -> NA, DUKES, MA (no such port)

## Species 

* Species have duplicate names with different formats
  * All species with NK or (NK) (Not known) in their name were formatted to NK SPECIESNAME
  * All species separated by commas (eg DRUM, BLACK) were formatted as BLACK DRUM
  * LOBSTER was replaced by AMERICAN LOBSTER
  * LOLIGO SQUID was replaced by LONGFIN SQUID
  * OTHER FISH have 2 codes (526,529)
* Several NESPP3 codes are catch alls. For example:
  * NS SQUIDS
    * 803: MKTNM = UNCLASSIFIED 
    * split landings 50/50 into atlantis groups ISQ/LSQ)
  * SKATES
    * 365: MKTNM = UNCLASSIFIED, SMALL (WINGS), LARGE (WINGS)
    * Split into SK,LSK,WSK (proportions?)
  * OTHER FISH
    * 526: MKTNM = HEADS, FOOD, SPAWN
    * 529: MKTNM = INDUSTRIAL
  * NK DOGFISH 
    * 350: MKTNM: ROUND, TAILS, DRESSED, FINS
  * NK SHARK
    * 359: UNCLASSIFIED DRESSED/ROUND, FINS DRIED/FRESH/FROZEN,GUTTED,CHUNKS
  * NK CLAM
    * 764:
  * HAKE MIX RED & WHITE
    * 155:
  * SILVER&OFFSHIREHAKE MIX
    * 507:
  
  
## GEARS
  
* Gear Codes have different Gear Categories assigned depending on VTR or DMIS data
  *   Use DMIS codes

