---
title: "Data Cleaning"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(atlantiscas)
```

The fishing data is a pulled from CAMS (1996-2021) data.

```{r sampleData, echo = F}
# read sample data and display it in a table
sampleData <- readRDS(here::here("data-raw/sampleFishingData.rds"))  |>  
  dplyr::select(Year,PORT,STATE,GEARCODE,GEARCAT, NESPP3,SPPNM)
DT::datatable(sampleData)
```


## Ports

* Port look up tables lack latitude and longitude
  * Use port, county, state to geocode lat and lon using open street maps
* Ports have duplicate entries with spelling typos
  * eg. harwichport, harwich port. These were hand fixed
  
  
All PORTID within the Atlantis domain were identified. Other PORTIDs were grouped into either Canadian ports (north of the atlantis domain) or southern ports (south of atlantis domain)

PORTID was used to match PORTNM,COUNTY,STATE from the CAMS_GARFO.CFG_PORT table.Port names were then used to find lat and lon values using `tidygeocoder` using open street maps. Many port names and/or counties had to be edited due to spelling errors, incorrect information, inconsistent formatting, or missing information. For example:

  * SALISBURY COVE, ME -> SALSBURY COVE, ME (spelling)
  * (TOWN OF) -> (inconsistent formatting -removed text "(TOWN OF)" from port name)
  * PERKINS COVE, CUMBERLAND, ME -> OGUNQUIT, YORK, ME (incorrect)
  * OTHER FAIRFIELD, FAIRFIELD, CT -> FAIRFIELD, FAIRFIELD, CT (missing port info)
  * COMBS CREEK, ST. MARYS, MD -> NA, ST.MARY'S, MD (not found on maps)
  * DUKES, DUKES, MA -> NA, DUKES, MA (no such port)
  * ORRS ISLAND, ME <- ORR'S ISLAND, ME
  * LONG ISLAND CUMBERLAND, ME <- LONG ISLAND, ME
  * STUEBEN, ME  <- STEUBEN, ME
  * WESTERLEY, RI <- WESTERLY, RI
  * BARINGTON, RI <- BARRINGTON, RI
  * DYERS BAY, ME <- STEUBEN, ME
  * MATHAIS POINT, VA <- DAHLGREN, VA
  * HAMPTON/SEABROOK, NH <- SEABROOK, NH
  * HARRIMANS POINT, ME <- BROOKLIN, ME
  * BROOKLIN, ME <- HANCOCK (COUNTY), ME
  * KINGS (COUNTY), NY <- BROOKLYN (COUNTY)
  
The following phrases were removed from PORTNM:

  * OTHER
  * COUNTY 
  * TOWN OF
  * LYNNHAVEN
  * CAPTREE
  * PORT
  * AQUINNAH
  
The following were removed from COUNTY

  * CITY OF
  * COUNTY (except from NEW LONDON COUNTY, NEW HAVEN COUNTY)
  * NOT-SPECIFIED

Lat and lons were found by matching the port data in the folloiing order

  * PORTNM, COUNTY, STATE
  * COUNTY, STATE
  * PORTNM, STATE
  
The remaining records in the data with ports that did not have a PORTID were then matched to existing/known ports (with PORTIDs) based on the recorded PORTLANDED, STATE values.

Some records (negligible) were discarded since there was no identifying port, or state associated with the catch)


## Species 

* Species have duplicate names with different formats
  * All species with NK or (NK) (Not known) in their name were formatted to NK SPECIESNAME
  * All species separated by commas (eg DRUM, BLACK) were formatted as BLACK DRUM
  * LOBSTER was replaced by AMERICAN LOBSTER
  * LOLIGO SQUID was replaced by LONGFIN SQUID
  * OTHER FISH have 2 codes (526,529)
  
* Several NESPP3 codes are catch alls. For example:
  * NS SQUIDS
    * 803: MKTNM = UNCLASSIFIED 
    * split landings 50/50 into atlantis groups ISQ/LSQ)
  * SKATES
    * 365: MKTNM = UNCLASSIFIED, SMALL (WINGS), LARGE (WINGS)
    * Split into SK,LSK,WSK (proportions?)
  * OTHER FISH
    * 526: MKTNM = HEADS, FOOD, SPAWN
    * 529: MKTNM = INDUSTRIAL
  * NK DOGFISH 
    * 350: MKTNM: ROUND, TAILS, DRESSED, FINS
  * NK SHARK
    * 359: UNCLASSIFIED DRESSED/ROUND, FINS DRIED/FRESH/FROZEN,GUTTED,CHUNKS
  * NK CLAM
    * 764:
  * HAKE MIX RED & WHITE
    * 155:
  * SILVER&OFFSHIREHAKE MIX
    * 507:
  
  
## GEARS
  
* Gear Codes have different Gear Categories assigned depending on VTR or DMIS data
  *   Use DMIS codes

