---
title: "Explore Fishing Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explore Fishing Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(atlantiscas)
library(magrittr)
```

```{r readdata, echo = F}
boundaryBoxes <- c(0,23:29)
lbstotons <- 2204.62

getData <- read_data()

allData <- getData$data |> 
  dplyr::mutate(Year = as.numeric(Year))
outsideData <- getData$outside

```

## Overview

A General overview of data. 
More detailed exploration based on Fleets, Ports, Species, can be found in other vignettes

## Data Sources and cleaning

Landings data by box at trip level provided by SSB for the communities at sea project

Example of data

```{r,sampledata, echo = F}
# read sample data and display it in a table
sampleData <- readRDS(here::here("data-raw/sampleFishingData.rds"))

DT::datatable(sampleData)
```

Explain the fields


## Plot Catch by Box {.tabset}

### NEUS

```{r plotlandingsbybox, echo = F}
# plot landings by box

datap <- allData |> 
  dplyr::group_by(Year,Area) |> 
  dplyr::summarise(tot = sum(InsideLANDED),
                   .groups="drop")  |> 
  tidyr::separate(Area,into = c("first","box"),sep="BOX_ID_")  |>  
  dplyr::mutate(box = as.numeric(box))  |> 
  dplyr::mutate(boundaryBox = (box %in% boundaryBoxes)) |> 
  dplyr::select(-first)
#  {. ->> d} %>%
  #dplyr::filter(box <= 22 & box > 0) %>%
p <- ggplot2::ggplot(datap) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=tot/lbstotons,color = boundaryBox)) +
  ggplot2::facet_wrap(ggplot2::vars(box),scales = "fixed")  +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) + 
  ggplot2::ggtitle("Total Landings by Atlantis Box")+
  ggplot2::ylab("Metric Tons")

suppressMessages(print(p))

```

### Outside


```{r plotlandingsbyboxoutside, echo = F}
# plot landings by box

datap <- outsideData |> 
  dplyr::group_by(Year,Area) |> 
  dplyr::summarise(tot = sum(InsideLANDED),
                   .groups="drop")  |> 
  tidyr::separate(Area,into = c("first","box"),sep="BOX_ID_")  |>  
  dplyr::mutate(box = as.numeric(box))  |> 
  dplyr::mutate(boundaryBox = (box %in% boundaryBoxes)) |> 
  dplyr::select(-first)
#  {. ->> d} %>%
  #dplyr::filter(box <= 22 & box > 0) %>%
p <- ggplot2::ggplot(datap) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=tot/lbstotons,color = boundaryBox)) +
  ggplot2::facet_wrap(ggplot2::vars(box),scales = "fixed")  +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) + 
  ggplot2::ggtitle("Total Landings by Atlantis Box")+
  ggplot2::ylab("Metric Tons")

suppressMessages(print(p))

```

## Plot catch by box totals {.tabset}

### Neus

```{r, plotcatchbox, echo = F}
  # plot catch by box
  p <- allData  |> 
    dplyr::group_by(Area)  |> 
    dplyr::summarise(tot = sum(InsideLANDED),
                     .groups="drop")  |> 
    tidyr::separate(Area,into = c("first","box"),sep="BOX_ID_") |> 
    dplyr::mutate(box = dplyr::case_when(as.numeric(box) < 10 ~ paste0("0",box),
                                         TRUE ~ box)) |> 
    dplyr::mutate(box = as.factor(box)) |> 
    dplyr::select(-first) |> 
    ggplot2::ggplot() +
    ggplot2::geom_col(ggplot2::aes(x=box,y=tot/(1000*lbstotons))) +
    ggplot2::ggtitle(paste("Total catch by Atlantis box from",paste(range(allData$Year),collapse = "-"))) +
    ggplot2::ylab("Metric tonnes (thousands)")+
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5))

print(p)

```

```{r plotboxprop,echo = F}
  # plot proportion of catch by box
  p <- allData %>%
    dplyr::group_by(Area) %>%
    dplyr::summarise(tot = sum(InsideLANDED),
                     .groups="drop") %>%
    tidyr::separate(.,Area,into = c("first","box"),sep="BOX_ID_") %>%
    dplyr::mutate(box = dplyr::case_when(as.numeric(box) < 10 ~ paste0("0",box),
                                         TRUE ~ box)) %>%
    dplyr::mutate(box = as.factor(box)) %>%
    dplyr::select(-first) %>%
    dplyr::mutate(prop = tot/sum(tot)) %>%
    ggplot2::ggplot(.) +
    ggplot2::geom_col(ggplot2::aes(x=box,y=prop)) +
    ggplot2::ggtitle(paste("Proportion of total catch by Atlantis box",paste(range(allData$Year),collapse = "-"))) +
    ggplot2::ylab("Proportion") +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5))

  print(p)
```

### Outside

```{r, plotcatchboxOutside, echo = F}
  # plot catch by box
  p <- outsideData  |> 
    dplyr::group_by(Area)  |> 
    dplyr::summarise(tot = sum(InsideLANDED),
                     .groups="drop")  |> 
    tidyr::separate(Area,into = c("first","box"),sep="BOX_ID_") |> 
    dplyr::mutate(box = dplyr::case_when(as.numeric(box) < 10 ~ paste0("0",box),
                                         TRUE ~ box)) |> 
    dplyr::mutate(box = as.factor(box)) |> 
    dplyr::select(-first) |> 
    ggplot2::ggplot() +
    ggplot2::geom_col(ggplot2::aes(x=box,y=tot/(1000*lbstotons))) +
    ggplot2::ggtitle(paste("Total catch by Atlantis box from",paste(range(allData$Year),collapse = "-"))) +
    ggplot2::ylab("Metric tonnes (thousands)")+
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5))

print(p)

```

```{r plotboxpropOutside,echo = F}
  # plot proportion of catch by box
  p <- outsideData %>%
    dplyr::group_by(Area) %>%
    dplyr::summarise(tot = sum(InsideLANDED),
                     .groups="drop") %>%
    tidyr::separate(.,Area,into = c("first","box"),sep="BOX_ID_") %>%
    dplyr::mutate(box = dplyr::case_when(as.numeric(box) < 10 ~ paste0("0",box),
                                         TRUE ~ box)) %>%
    dplyr::mutate(box = as.factor(box)) %>%
    dplyr::select(-first) %>%
    dplyr::mutate(prop = tot/sum(tot)) %>%
    ggplot2::ggplot(.) +
    ggplot2::geom_col(ggplot2::aes(x=box,y=prop)) +
    ggplot2::ggtitle(paste("Proportion of total catch by Atlantis box",paste(range(allData$Year),collapse = "-"))) +
    ggplot2::ylab("Proportion") +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5))

  print(p)
```




## Landings in Boundary boxes {.tabset}

Atlantis boundary boxes are not modelled like other boxes so we need to know how much of the landings come from these boxes

### Neus

```{r boundarybox, echo =F}

 # total landings by box (metric tons)
  totData <- allData %>%
    dplyr::group_by(Area) %>%
    dplyr::summarise(tot = sum(InsideLANDED)/lbstotons,
                     .groups="drop") %>%
    tidyr::separate(.,Area,into = c("first","box"),sep="BOX_ID_") %>%
    dplyr::mutate(box = dplyr::case_when(as.numeric(box) < 10 ~ paste0("0",box),
                                         TRUE ~ box)) %>%
    dplyr::mutate(box = as.factor(box)) %>%
    dplyr::select(-first)

  # landings in boundary boxes
  bbtot <- totData %>%
    dplyr::filter(box %in% boundaryBoxes) %>%
    dplyr::pull(tot) %>%
    sum()
  # proportion of landings in boundary boxes
  bbprop <- bbtot/(sum(totData$tot))
```

Total landings in boundary boxes over the period 1996-2021 is `r bbtot` which accounts for `r 100*bbprop` percent of the total landings over this period



### Outside

```{r boundaryboxOutside, echo =F}

 # total landings by box (metric tons)
  totData <- outsideData %>%
    dplyr::group_by(Area) %>%
    dplyr::summarise(tot = sum(InsideLANDED)/lbstotons,
                     .groups="drop") %>%
    tidyr::separate(.,Area,into = c("first","box"),sep="BOX_ID_") %>%
    dplyr::mutate(box = dplyr::case_when(as.numeric(box) < 10 ~ paste0("0",box),
                                         TRUE ~ box)) %>%
    dplyr::mutate(box = as.factor(box)) %>%
    dplyr::select(-first)

  # landings in boundary boxes
  bbtot <- totData %>%
    dplyr::filter(box %in% boundaryBoxes) %>%
    dplyr::pull(tot) %>%
    sum()
  # proportion of landings in boundary boxes
  bbprop <- bbtot/(sum(totData$tot))
```

Total landings in boundary boxes over the period 1996-2021 is `r bbtot` which accounts for `r 100*bbprop` percent of the total landings (by trips from States outside of NEUS) over this period


## Port level

There are `r length(unique(allData$newport)) + length(unique(outsideData$PORTID))` unique ports along the US eastern seaboard in which fish caught in the NEUS region are landed. `r length(unique(allData$newport))` are within NEUS


The Main ports in terms of landings (1000's metric tons). Only states within the model are considered

```{r mainports, echo=FALSE}


  ## box catch  at port level
  totportboxdata <- allData %>%
    dplyr::group_by(newport,STATEABB,Area) %>%
    dplyr::summarise(tot = sum(InsideLANDED)/lbstotons,
                     .groups="drop") %>%
    tidyr::separate(.,Area,into = c("first","box"),sep="BOX_ID_") %>%
    dplyr::mutate(box = dplyr::case_when(as.numeric(box) < 10 ~ paste0("0",box),
                                         TRUE ~ box)) %>%
    dplyr::mutate(box = as.factor(box)) %>%
    dplyr::select(-first)

  #landings by port, sum over boxes
  totportdata <- totportboxdata %>%
    dplyr::group_by(newport,STATEABB) %>%
    dplyr::summarise(tot = sum(tot),
                     .groups="drop")
  # Main ports for landings
  mainports <- totportdata %>%
    dplyr::arrange(desc(tot))
  
  DT::datatable(mainports)
  
  
```
## Landings by Ports and State

Only states within the model are considered
Top 5 ports in each state based on landings (1000's metric tons)

```{r landportstate, echo = F}

  ## top 5 ports in each state
  landingstop5 <- totportdata |>
    dplyr::group_by(STATEABB) |> 
    dplyr::slice_max(order_by = tot, n = 5)
  

  DT::datatable(landingstop5)


```




## Ports by State

Lighly shaded bars indicate ports that lie outside of the NEUS footprint.
Note: CN = Canadian ports

```{r portsbystateall, echo =FALSE}

  # plot number of ports by state
  portlevel <- c("CN","ME","NH","MA","RI","CT","NY","NJ","PA","DE","MD","VA","NC","SC","GA","FL")
  highlightarea <- c("ME","NH","MA","RI","CT","NY","NJ","PA","DE","MD","VA")
  ## box catch  at port level
  outsideData2 <- outsideData |>
    dplyr::mutate(newport=PORTID,
                  newcounty=NA,
                  lat=NA,
                  lon=NA) |> 
    dplyr::relocate(names(allData))
  
  
  combinedData <- rbind(allData,outsideData2)

  totportboxdata <- combinedData %>%
    dplyr::group_by(newport,STATEABB,Area) %>%
    dplyr::summarise(tot = sum(InsideLANDED)/lbstotons,
                     .groups="drop") %>%
    tidyr::separate(.,Area,into = c("first","box"),sep="BOX_ID_") %>%
    dplyr::mutate(box = dplyr::case_when(as.numeric(box) < 10 ~ paste0("0",box),
                                         TRUE ~ box)) %>%
    dplyr::mutate(box = as.factor(box)) %>%
    dplyr::select(-first)

  #landings by port, sum over boxes
  totportdata <- totportboxdata %>%
    dplyr::group_by(newport,STATEABB) %>%
    dplyr::summarise(tot = sum(tot),
                     .groups="drop")
  # plot number of ports by state

  totportdata %>%
    dplyr::group_by(STATEABB) %>%
    dplyr::mutate(STATEABB = factor(STATEABB,level=portlevel)) %>%
    dplyr::summarise(numports = length(unique(newport))) %>%
    dplyr::filter(!is.na(STATEABB)) %>%
    dplyr::arrange(desc(numports)) %>%
    ggplot2::ggplot(.) +
    ggplot2::geom_col(ggplot2::aes(x=STATEABB,y = numports)) +
    ggplot2::ylab("Number of ports") +
    gghighlight::gghighlight(STATEABB %in% highlightarea)
```


```{r portdata, echo = F}


  #landings by port in boundary boxes
  bbportdata <- totportboxdata %>%
    dplyr::filter(box %in% boundaryBoxes) %>%
    dplyr::group_by(newport,STATEABB) %>%
    dplyr::summarise(tot = sum(tot),
                     .groups="drop")

  
```

Are the fish landed from the boundary boxes distributed equally among ports or are some ports heavily reliant from landings in these boxes

```{r portboundary, echo = F}
  # proportion of port landing from boundary box
  pp <- totportdata %>%
    dplyr::left_join(.,bbportdata,by = c("newport","STATEABB")) %>%
    dplyr::mutate(prop = dplyr::case_when(is.na(tot.y) ~ 0,
                                          TRUE ~ tot.y/tot.x)) %>%
    dplyr::arrange(desc(prop))

 DT::datatable(pp)

```


## Landings by State

```{r stateland, echo = F}
  allData %>%
    dplyr::group_by(Year,STATEABB) %>%
    dplyr::summarise(tot = sum(InsideLANDED)/lbstotons,.groups="drop") %>%
    ggplot2::ggplot(.) +
    ggplot2::geom_col(ggplot2::aes(x=as.numeric(Year),y=tot)) +

    #ggplot2::geom_line(ggplot2::aes(x=as.numeric(Year),y=tot)) +
    ggplot2::facet_wrap(ggplot2::vars(STATEABB),scales = "free_y") +
    ggplot2::theme(axis.text.x=ggplot2::element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ggplot2::ylab("Metric Tons (1000s)") + 
  ggplot2::xlab("") + 
  ggplot2::ggtitle("Landings by State")


```

Missing State/port designations appear in data. This results in a number if trip not being able to be attributed to a port. Table shows number of trips with missing port/state information

```{r missingports,echo = F}
 allData %>%
    dplyr::filter(is.na(STATEABB)) %>%
    dplyr::group_by(Year) %>%
    dplyr::summarise(numTrips=dplyr::n()) %>%
  DT::datatable()
```

## Number of Trips

A trip is identified by unique TRIPID.

```{r numrips, echo = F}
p <- allData %>%
    dplyr::mutate(Year = as.numeric(Year),
                  TRIPID = as.numeric(TRIPID)) %>%
    dplyr::group_by(Year) %>%
    dplyr::summarise(numTrip = length(unique(TRIPID))) %>%
    ggplot2::ggplot(.) +
    ggplot2::geom_col(ggplot2::aes(x=Year,y=numTrip)) + 
  ggplot2::ylab("Number of Trips") + 
  ggplot2::xlab("") 
print(p)

```


## Gear categories

```{r gearcodes ,echo = F}
# cams data
allData %>%
  dplyr::distinct(GEARCODE, GEARCAT) %>%
  { . ->> gearTable} %>%
  DT::datatable()

```



There are no GEARCODEs that have missing GEARCATs. The distinct Gear categories DMIS are :
```{r dmiscats, echo =FALSE}
allData %>% 
  dplyr::distinct(GEARCAT) %>%
  dplyr::arrange(GEARCAT) %>%
  DT::datatable(options=list(lengthMenu=list(c("20")),paging=T))
  
```

## Landings by GEARCAT

Look at total landings by gear type/and box

```{r gearcat, echo = F}
allData %>%
  dplyr::group_by(Year,STATEABB,GEARCAT) %>%
  dplyr::summarize(mt = sum(InsideLANDED)/lbstotons,
                   .groups="drop") %>% 
  {. ->> gearStateData} %>%
  ggplot2::ggplot(.) +
  ggplot2::geom_line(ggplot2::aes(x=as.numeric(Year),y=mt,color=STATEABB)) +
  ggplot2::facet_wrap(ggplot2::vars(GEARCAT)) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle=90))
  

gearStateData %>% 
  dplyr::group_by(STATEABB) %>%
  dplyr::summarise(mt = ceiling(sum(mt)),.groups = "drop") %>% 
  dplyr::arrange(desc(mt)) %>% 
  DT::datatable()


gearStateData %>% 
  dplyr::group_by(GEARCAT) %>%
  dplyr::summarise(mt = ceiling(sum(mt)),.groups = "drop") %>% 
  dplyr::arrange(desc(mt)) %>% 
  DT::datatable()

```



## Species categories

There are some species without an atlantis code assigned. Some of these are "catch all" species codes that will need to be resolved. Eg. SKATES, NK SHARK, NK CLAM, OTHER FISH. Others can be ignored

```{r speciesMissingCodes, echo = F}

p<- allData |> 
  dplyr::filter(is.na(Code)) |>
  dplyr::group_by(Year,NESPP3,SPPNM) |>
    dplyr::summarise(lbs = sum(InsideLANDED),
                     .groups = "drop") |>
    ggplot2::ggplot() +
    ggplot2::geom_line(ggplot2::aes(x=as.numeric(Year),y=lbs)) +
    ggplot2::facet_wrap(~as.factor(SPPNM),scales = "free_y")

suppressWarnings(print(p))

```

### Skates

All species identified as a SKATE in the data
```{r skates, echo = F}
allData |>
    dplyr::filter(grepl("SKATE",SPPNM)) |>
    dplyr::group_by(Year,NESPP3,SPPNM) |>
    dplyr::summarise(lbs = sum(InsideLANDED),
                     .groups = "drop") |>
    ggplot2::ggplot() +
    ggplot2::geom_line(ggplot2::aes(x=as.numeric(Year),y = lbs)) +
    ggplot2::facet_wrap(~SPPNM) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```

### Squid

All species identified as a SQUID in the data

```{r squid, echo = F}
allData |>
    dplyr::filter(grepl("SQUID",SPPNM)) |>
    dplyr::group_by(Year,NESPP3,SPPNM) |>
    dplyr::summarise(lbs = sum(InsideLANDED),
                     .groups = "drop") |>
    ggplot2::ggplot() +
    ggplot2::geom_line(ggplot2::aes(x=as.numeric(Year),y = lbs)) +
    ggplot2::facet_wrap(~SPPNM) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```

### Shark

All species identified as a SHARK in the data

```{r shark, echo = F}
allData |>
    dplyr::filter(grepl("SHARK",SPPNM)) |>
    dplyr::group_by(Year,NESPP3,SPPNM) |>
    dplyr::summarise(lbs = sum(InsideLANDED),
                     .groups = "drop") |>
    ggplot2::ggplot() +
    ggplot2::geom_line(ggplot2::aes(x=as.numeric(Year),y = lbs)) +
    ggplot2::facet_wrap(~SPPNM) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```


### Dogfish

All species identified as a DOGFISH in the data

```{r dogfish, echo = F}
allData |>
    dplyr::filter(grepl("DOGFISH",SPPNM)) |>
    dplyr::group_by(Year,NESPP3,SPPNM) |>
    dplyr::summarise(lbs = sum(InsideLANDED),
                     .groups = "drop") |>
    ggplot2::ggplot() +
    ggplot2::geom_line(ggplot2::aes(x=as.numeric(Year),y = lbs)) +
    ggplot2::facet_wrap(~SPPNM) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```

### Clam

All species identified as a CLAM in the data

```{r clam, echo = F}
allData |>
    dplyr::filter(grepl("CLAM",SPPNM)) |>
    dplyr::group_by(Year,NESPP3,SPPNM) |>
    dplyr::summarise(lbs = sum(InsideLANDED),
                     .groups = "drop") |>
    ggplot2::ggplot() +
    ggplot2::geom_line(ggplot2::aes(x=as.numeric(Year),y = lbs)) +
    ggplot2::facet_wrap(~SPPNM) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```




```{r speciescat, echo = F}

neusGroups <- readr::read_csv(url("https://raw.githubusercontent.com/NOAA-EDAB/neus-atlantis/master/data/functionalGroupNames.csv"),col_types = readr::cols(Code = readr::col_character(),
                                                                                                                                                            Functional_Group = readr::col_character(),
                                                                                                                                                            Scientific_Name = readr::col_character(),
                                                                                                                                                            SVSPP= readr::col_double(),
                                                                                                                                                            NESPP3 = readr::col_double(),
                                                                                                                                                            Species_Itis = readr::col_double(),
                                                                                                                                                            isFishedSpecies = readr::col_logical())) %>% 
  dplyr::select(Code, Species, NESPP3)

# write out list of species with NA codes in Atlantis functional group names
if(0) {
allData %>%
  dplyr::left_join(.,neusGroups, by = "NESPP3") %>%
  dplyr::group_by(SPPNM,NESPP3,Code,Species) %>%
  dplyr::summarise(mt = sum(InsideLANDED)/lbstotons) %>%
  dplyr::filter(is.na(Code)) %>% 
  dplyr::arrange(desc(mt)) %>% 
  readr::write_csv(.,here::here("data-raw/NASpecies.csv"))
}

# dd %>%
#   dplyr::select(SPPNM,NESPP3,Code,Species,InsideLANDED) %>%
#   dplyr::filter(is.na(Code)) %>%
#   dplyr::group_by(SPPNM) %>%
#   dplyr::summarise(tot = sum(InsideLANDED))




```

