---
title: "Explore Fishing Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explore Fishing Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(atlantiscas)
library(magrittr)
```

```{r readdata, echo = F}
boundaryBoxes <- c(0,23:29)
lbstotons <- 2204.62

data <- read_data()$data

```
## Data Sources and cleaning

Landings data by box at trip level provided by SSB for the communities at sea project

Example of data

```{r,sampledata, echo = F}
# read sample data and display it in a table
sampleData <- readRDS(here::here("data-raw/sampleFishingData.rds"))

DT::datatable(sampleData)
```

Explain the fields


### Plot Catch by Box


```{r plotlandingsbybox, echo = F}
# plot landings by box
p <- data %>%
  dplyr::group_by(Year,Area) %>% 
  dplyr::summarise(tot = sum(InsideLANDED),.groups="drop") %>%
  tidyr::separate(.,Area,into = c("first","box"),sep="BOX_ID_") %>% 
  dplyr::mutate(box = as.numeric(box)) %>%
  dplyr::mutate(boundaryBox = (box %in% boundaryBoxes)) %>% 
  dplyr::select(-first) %>%
  {. ->> d} %>%
  #dplyr::filter(box <= 22 & box > 0) %>%
  ggplot2::ggplot(.) +
  ggplot2::geom_line(ggplot2::aes(x=as.numeric(Year),y=tot/lbstotons,color = boundaryBox)) +
  ggplot2::facet_wrap(ggplot2::vars(box),scales = "fixed")  +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) + 
  ggplot2::ggtitle("Total Landings by Atlantis Box")+
  ggplot2::ylab("Metric Tons")

suppressMessages(print(p))

```
```{r, plotcatchbox, echo = F}
  # plot catch by box
  p <- data %>%
    dplyr::group_by(Area) %>%
    dplyr::summarise(tot = sum(InsideLANDED),
                     .groups="drop") %>%
    tidyr::separate(.,Area,into = c("first","box"),sep="BOX_ID_") %>%
    dplyr::mutate(box = dplyr::case_when(as.numeric(box) < 10 ~ paste0("0",box),
                                         TRUE ~ box)) %>%
    dplyr::mutate(box = as.factor(box)) %>%
    dplyr::select(-first) %>%
    ggplot2::ggplot(.) +
    ggplot2::geom_col(ggplot2::aes(x=box,y=tot/(1000*lbstotons))) +
    ggplot2::ggtitle(paste("Total catch by Atlantis box from",paste(range(data$Year),collapse = "-"))) +
    ggplot2::ylab("Metric tonnes (thousands)")+
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5))

print(p)

```

```{r plotboxprop,echo = F}
  # plot proportion of catch by box
  p <- data %>%
    dplyr::group_by(Area) %>%
    dplyr::summarise(tot = sum(InsideLANDED),
                     .groups="drop") %>%
    tidyr::separate(.,Area,into = c("first","box"),sep="BOX_ID_") %>%
    dplyr::mutate(box = dplyr::case_when(as.numeric(box) < 10 ~ paste0("0",box),
                                         TRUE ~ box)) %>%
    dplyr::mutate(box = as.factor(box)) %>%
    dplyr::select(-first) %>%
    dplyr::mutate(prop = tot/sum(tot)) %>%
    ggplot2::ggplot(.) +
    ggplot2::geom_col(ggplot2::aes(x=box,y=prop)) +
    ggplot2::ggtitle(paste("Proportion of total catch by Atlantis box",paste(range(data$Year),collapse = "-"))) +
    ggplot2::ylab("Proportion") +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5))

  print(p)
```


### Landings in Boundary boxes

Atlantis boundary boxes are not modelled like other boxes so we need to know how much of the landings come from these boxes

```{r boundarybox, echo =F}

 # total landings by box (metric tons)
  totData <- data %>%
    dplyr::group_by(Area) %>%
    dplyr::summarise(tot = sum(InsideLANDED)/lbstotons,
                     .groups="drop") %>%
    tidyr::separate(.,Area,into = c("first","box"),sep="BOX_ID_") %>%
    dplyr::mutate(box = dplyr::case_when(as.numeric(box) < 10 ~ paste0("0",box),
                                         TRUE ~ box)) %>%
    dplyr::mutate(box = as.factor(box)) %>%
    dplyr::select(-first)

  # landings in boundary boxes
  bbtot <- totData %>%
    dplyr::filter(box %in% boundaryBoxes) %>%
    dplyr::pull(tot) %>%
    sum()
  # proportion of landings in boundary boxes
  bbprop <- bbtot/(sum(totData$tot))
```

Total landings in boundary boxes over the period 1996-2021 is `r bbtot` which accounts for `r 100*bbprop` percent of the total landings over this period

## Port level

There are `r length(unique(data$PORTLANDED))` unique ports along the US eastern seaboard in which fish caught in the NEUS region are landed


The Main ports in terms of landings (1000's metric tons)

```{r mainports, echo=FALSE}


  ## box catch  at port level
  totportboxdata <- data %>%
    dplyr::group_by(PORTLANDED,STATE,Area) %>%
    dplyr::summarise(tot = sum(InsideLANDED)/lbstotons,
                     .groups="drop") %>%
    tidyr::separate(.,Area,into = c("first","box"),sep="BOX_ID_") %>%
    dplyr::mutate(box = dplyr::case_when(as.numeric(box) < 10 ~ paste0("0",box),
                                         TRUE ~ box)) %>%
    dplyr::mutate(box = as.factor(box)) %>%
    dplyr::select(-first)

  #landings by port, sum over boxes
  totportdata <- totportboxdata %>%
    dplyr::group_by(PORTLANDED,STATE) %>%
    dplyr::summarise(tot = sum(tot),
                     .groups="drop")
  # Main ports for landings
  mainports <- totportdata %>%
    dplyr::arrange(desc(tot))
  
  DT::datatable(mainports)
  
  
```
## Landings by Ports and State

Top 5 ports in each state based on landings (1000's metric tons)

```{r landportstate, echo = F}

  ## top 5 ports in each state
  landingstop5 <- totportdata |>
    dplyr::group_by(STATE) |> 
    dplyr::slice_max(order_by = tot, n = 5)
  

  DT::datatable(landingstop5)


```




## Ports by State

Lighly shaded bars indicate ports that lie outside of the NEUS footprint.
Note: CN = Canadian ports

```{r portsbystate, echo=F}

  # plot number of ports by state
  portlevel <- c("CN","ME","NH","MA","RI","CT","NY","NJ","PA","DE","MD","VA","NC","SC","GA","FL")
  highlightarea <- c("ME","NH","MA","RI","CT","NY","NJ","PA","DE","MD","VA")
  totportdata %>%
    dplyr::group_by(STATE) %>%
    dplyr::mutate(STATE = factor(STATE,level=portlevel)) %>%
    dplyr::summarise(numports = length(unique(PORTLANDED))) %>%
    dplyr::filter(!is.na(STATE)) %>%
    dplyr::arrange(desc(numports)) %>%
    {. ->> portd} %>%
    ggplot2::ggplot(.) +
    ggplot2::geom_col(ggplot2::aes(x=STATE,y = numports)) +
    ggplot2::ylab("Number of ports") +
    gghighlight::gghighlight(STATE %in% highlightarea)

```

```{r portdata, echo = F}


  #landings by port in boundary boxes
  bbportdata <- totportboxdata %>%
    dplyr::filter(box %in% boundaryBoxes) %>%
    dplyr::group_by(PORTLANDED,STATE) %>%
    dplyr::summarise(tot = sum(tot),
                     .groups="drop")

  
```

Are the fish landed from the boundary boxes distributed equally among ports or are some ports heavily reliant from landings in these boxes

```{r portboundary, echo = F}
  # proportion of port landing from boundary box
  pp <- totportdata %>%
    dplyr::left_join(.,bbportdata,by = c("PORTLANDED","STATE")) %>%
    dplyr::mutate(prop = dplyr::case_when(is.na(tot.y) ~ 0,
                                          TRUE ~ tot.y/tot.x)) %>%
    dplyr::arrange(desc(prop))

 DT::datatable(pp)

```


## Landings by State

```{r stateland, echo = F}
  data %>%
    dplyr::group_by(Year,STATE) %>%
    dplyr::summarise(tot = sum(InsideLANDED)/lbstotons,.groups="drop") %>%
    ggplot2::ggplot(.) +
    ggplot2::geom_col(ggplot2::aes(x=as.numeric(Year),y=tot)) +

    #ggplot2::geom_line(ggplot2::aes(x=as.numeric(Year),y=tot)) +
    ggplot2::facet_wrap(ggplot2::vars(STATE),scales = "free_y") +
    ggplot2::theme(axis.text.x=ggplot2::element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ggplot2::ylab("Metric Tons (1000s)") + 
  ggplot2::xlab("") + 
  ggplot2::ggtitle("Landings by State")


```

Missing State/port designations appear in data. This results in a number if trip not being able to be attributed to a port. Table shows number of trips with missing port/state information

```{r missingports,echo = F}
 data %>%
    dplyr::filter(is.na(STATE)) %>%
    dplyr::group_by(Year) %>%
    dplyr::summarise(numTrips=dplyr::n()) %>%
  DT::datatable()
```

## Number of Trips

A trip is identified by unique TRIPID.

```{r numrips, echo = F}
p <- data %>%
    dplyr::mutate(Year = as.numeric(Year),
                  TRIPID = as.numeric(TRIPID)) %>%
    dplyr::group_by(Year) %>%
    dplyr::summarise(numTrip = length(unique(TRIPID))) %>%
    ggplot2::ggplot(.) +
    ggplot2::geom_col(ggplot2::aes(x=Year,y=numTrip)) + 
  ggplot2::ylab("Number of Trips") + 
  ggplot2::xlab("") 
print(p)

```


## Gear categories

```{r gearcodes ,echo = F}
# vtr data
d1 <- data %>%
  dplyr::filter(Year < 2008) %>%
  dplyr::distinct(GEARCODE, GEARCAT) 

# dmis
d2 <- data %>% 
  dplyr::filter(Year >=2008) %>%
  dplyr::distinct(GEARCODE, GEARCAT) 


d1 %>%
  dplyr::full_join(.,d2, by = c("GEARCODE")) %>%
  dplyr::rename(VTR = GEARCAT.x,
                DMIS = GEARCAT.y) %>%
  dplyr::arrange(DMIS) %>%
  { . ->> gearTable} %>%
  DT::datatable()
  
```



There are no GEARCODEs in DMIS that have missing GEARCATs. VTR and DMIS GEARCAT descriptions match. The distinct Gear categories from DMIS are :
```{r dmiscats, echo =FALSE}
data %>% 
  dplyr::filter(Year >=2008) %>%
  dplyr::distinct(GEARCAT) %>%
  dplyr::arrange(GEARCAT) %>%
  DT::datatable(options=list(lengthMenu=list(c("20")),paging=T))
  
```

## Landings by GEARCAT

Map all DMIS codes/categories to VTR codes.
Look at total landings by gear type/and box

```{r gearcat, echo = F}
data %>%
  dplyr::left_join(.,gearTable,by="GEARCODE") %>%
  dplyr::select(-GEARCAT,-VTR) %>%
  dplyr::rename(GEARCAT = DMIS) %>%
  dplyr::group_by(Year,STATE,GEARCAT) %>%
  dplyr::summarize(mt = sum(InsideLANDED)/lbstotons,groups="drop") %>% 
  {. ->> gearStateData} %>%
  ggplot2::ggplot(.) +
  ggplot2::geom_line(ggplot2::aes(x=as.numeric(Year),y=mt,color=STATE)) +
  ggplot2::facet_wrap(ggplot2::vars(GEARCAT)) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle=90))
  

gearStateData %>% 
  dplyr::group_by(STATE) %>%
  dplyr::summarise(mt = ceiling(sum(mt)),.groups = "drop") %>% 
  dplyr::arrange(desc(mt)) %>% 
  DT::datatable()


gearStateData %>% 
  dplyr::group_by(GEARCAT) %>%
  dplyr::summarise(mt = ceiling(sum(mt)),.groups = "drop") %>% 
  dplyr::arrange(desc(mt)) %>% 
  DT::datatable()

```



## Species categories


```{r speciescat, echo = F}

neusGroups <- readr::read_csv(url("https://raw.githubusercontent.com/NOAA-EDAB/neus-atlantis/master/data/functionalGroupNames.csv"),col_types = readr::cols(Code = readr::col_character(),
                                                                                                                                                            Functional_Group = readr::col_character(),
                                                                                                                                                            Scientific_Name = readr::col_character(),
                                                                                                                                                            SVSPP= readr::col_double(),
                                                                                                                                                            NESPP3 = readr::col_double(),
                                                                                                                                                            Species_Itis = readr::col_double(),
                                                                                                                                                            isFishedSpecies = readr::col_logical())) %>% 
  dplyr::select(Code, Species, NESPP3)

# write out list of species with NA codes in Atlantis functional group names
if(0) {
data %>%
  dplyr::left_join(.,neusGroups, by = "NESPP3") %>%
  dplyr::group_by(SPPNM,NESPP3,Code,Species) %>%
  dplyr::summarise(mt = sum(InsideLANDED)/lbstotons) %>%
  dplyr::filter(is.na(Code)) %>% 
  dplyr::arrange(desc(mt)) %>% 
  readr::write_csv(.,here::here("data-raw/NASpecies.csv"))
}

# dd %>%
#   dplyr::select(SPPNM,NESPP3,Code,Species,InsideLANDED) %>%
#   dplyr::filter(is.na(Code)) %>%
#   dplyr::group_by(SPPNM) %>%
#   dplyr::summarise(tot = sum(InsideLANDED))




```

