---
title: "NE Groundfish fleet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NE Groundfish fleet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(atlantiscas)
library(ncdf4)
library(sf)
```


## Species

The [NE multispecies (Groundfish) FMP](https://www.nefmc.org/management-plans/northeast-multispecies) comprises the following species:

* Atlantic Cod (COD)
* Haddock (HAD)
* Yellowtail Flounder (YTF)
* Pollock (POL)
* American Plaice (PLA)
* Witch Flounder (WTF)
* White Hake (WHK)
* Winter Flounder (WIF)
* Redfish (RED)
* Atlantic Halibut (HAL)

Non target species

* Windowpane flounder (WPF)
* Ocean Pout (OPT)
* Atlantic Wolffish (WOL)

These GEARCAT(s) associated with groundfish species are:

* Bottom Trawl
* Sink Gillnet

** Still need to separate some species (skates, sharks) **


```{r groundfishdata, echo = F}
gf <- readRDS(here::here("data/NEGroundfishDataCAMS.rds")) |> 
  dplyr::mutate(Year = as.integer(Year))

gfCodes <- c("COD","HAD","YTF","POL","PLA","WTF","WHK","WIF","RED","HAL","WPF","OPT","WOL")

gf |> dplyr::distinct(GEARCAT,GEARCODE) |> 
  dplyr::arrange(GEARCAT) |> 
  DT::datatable()
```

## Landings

The landings of trips with associated bycatch

```{r plotland1, echo = F}
landings <- gf |> 
  dplyr::group_by(Year,Functional_Group) |> 
  dplyr::summarise(lbs = sum(InsideLANDED),
                   .groups = "drop")
                     
speciesLanded <- landings |> 
  dplyr::group_by(Functional_Group) |> 
  dplyr::summarise(nyrs = dplyr::n(),
                   .groups = "drop") |> 
  dplyr::filter(nyrs > 1) |> 
  dplyr::pull(Functional_Group)

landings |> 
  dplyr::filter(Functional_Group %in% speciesLanded) |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=lbs)) +
  ggplot2::facet_wrap(~Functional_Group)

```

## Total groundfish landings for species in FMP

```{r plotcompare, echo = F}
landings <- gf |> 
  dplyr::filter(Code %in% gfCodes) |>  
  dplyr::group_by(Year,Code) |> 
  dplyr::summarise(lbs = sum(InsideLANDED),
                   .groups = "drop")

landings |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=lbs)) +
  ggplot2::facet_wrap(~Code)
```

## Total groundfish landings for bycatch species

```{r plotcompare2, echo = F}
landings <- gf |> 
  dplyr::filter(!(Code %in% gfCodes)) |>  
  dplyr::group_by(Year,Code) |> 
  dplyr::summarise(lbs = sum(InsideLANDED),
                   .groups = "drop")

landings |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=lbs)) +
  ggplot2::facet_wrap(~Code)
```

## Landings of groundfish by species

```{r plotland2, echo =F}

gf |> 
  dplyr::group_by(Functional_Group,Code) |> 
  dplyr::summarise(millionlbs = sum(InsideLANDED)/1000000,
                   .groups = "drop") |> 
  dplyr::arrange(desc(millionlbs)) |> 
  DT::datatable(options = list(pageLength = 10,
                               order = list(list(0, 'asc'))
          ))
```


## Landings of Groundfish by state

```{r plotland3, echo = F}
landings <- gf |> 
  dplyr::filter(Code %in% gfCodes) |>  
  dplyr::group_by(Year,STATEABB) |> 
  dplyr::summarise(lbs = sum(InsideLANDED),
                   .groups = "drop")
                     
landings |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=lbs)) +
  ggplot2::facet_wrap(~STATEABB)

```

## Landings of groundfish by port (species in FMP)

```{r plotland4, echo=F}
landings <- gf |> 
  dplyr::filter(Code %in% gfCodes) |>  
  dplyr::group_by(Year,STATEABB,newport) |> 
  dplyr::summarise(lbs = sum(InsideLANDED),
                   .groups = "drop")

ports10 <- landings |> 
  dplyr::group_by(newport,STATEABB) |> 
  dplyr::summarise(lbs = sum(lbs),
                   .groups = "drop") |> 
  dplyr::arrange(desc(lbs)) |> 
  dplyr::slice(1:10) |> 
  dplyr::pull(newport)


landings |> 
  dplyr::filter(newport %in% ports10) |> 
  dplyr::group_by(Year,newport) |> 
  dplyr::summarise(lbs = sum(lbs),
                   .groups="drop") |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=lbs)) +
  ggplot2::facet_wrap(~newport)

```


## Landings by Atlantis Box
compared to model output from v2.2.0

```{r plotland5, echo=F}
landings <- gf |> 
  dplyr::filter(Code %in% gfCodes) |>  
  dplyr::group_by(Year,Area) |> 
  dplyr::summarise(lbs = sum(InsideLANDED),
                   .groups = "drop") |>
  tidyr::separate(col=Area,into = c("text","Box"),sep = "ID_") |> 
  dplyr::mutate(Box = as.numeric(Box)) |> 
  dplyr::mutate(mt = lbs*0.000453592) |> 
  dplyr::select(-text,-lbs) |> 
  dplyr::mutate(source = "CAMS")

# Compare to atlantis

# 
# x.nc = nc_open(here::here("data-raw/atlantis",'neus_outputCATCH.nc'))
# varname = names(x.nc$var)
# vn <- grep('SCA_Catch_FC1',varname,value = T)
# 
# dd <- ncvar_get(x.nc,"SCA_Catch_FC1",collapse_degen = F)
# nc_close(x.nc)
# 
# rownames(dd) <- 1:nrow(dd)
# colnames(dd) <- 1:ncol(dd)
# 
# d2 <- as.data.frame(dd) |>
#   tibble::rownames_to_column(var="Box") |>
#   dplyr::mutate(Box = as.integer(Box) -1) |>
#   tidyr::pivot_longer(cols=-Box,names_to = "t",values_to = "value") |>
#   dplyr::mutate(t = as.integer(t),
#                 Year = 1963+ceiling(t/5)) |> 
#   dplyr::group_by(Box,Year) |> 
#   dplyr::summarise(mt = mean(value),
#                    .groups="drop") |> 
#   dplyr::relocate(Year,Box,mt) |> 
#   dplyr::mutate(source = "atlantis")
# 
# dataF <- rbind(d2,landings) |> 
#   dplyr::filter(Year > 1995)
# 

dataF <- landings
dataF |>
  ggplot2::ggplot() +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=mt,color=source)) +
  ggplot2::facet_wrap(~Box) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90))

```

## Landings (data) by box spatially

```{r plotland6, echo =F}
dataBox <- dataF |> 
  dplyr::filter(source == "CAMS") |> 
  dplyr::group_by(Box) |> 
  dplyr::summarise(mt = sum(mt))

# select boxes which account fo 95% of landings
landings <- dataBox |> 
  dplyr::arrange(desc(mt)) |> 
  dplyr::mutate(cumsum = cumsum(mt), prop = cumsum/sum(mt)) |> 
  dplyr::filter(prop <= 0.95)

NEFSCspatial::Neus_atlantis |> 
  dplyr::left_join(landings,by = c("BOX_ID"="Box"))  |> 
  ggplot2::ggplot() +
  ggplot2::geom_sf(ggplot2::aes(fill = mt)) +
  ggplot2::geom_sf_text(ggplot2::aes(label = BOX_ID),
                        position = ggplot2::position_dodge(width=0.9),  size=3) 

```

Boxes that contribute 95% of landings `r landings |> dplyr::pull(Box) |> sort()`


## Landings (data) by box spatially for main ports

```{r plotland7, echo =F}

# pick out 3 busiest ports based on landings
ports <- gf |> 
  dplyr::filter(Code %in% gfCodes) |> 
  dplyr::group_by(newport,PORTID,lat,lon) |> 
  dplyr::summarise(lb = sum(InsideLANDED),
                   .groups = "drop") |> 
  dplyr::arrange(desc(lb)) |> 
  dplyr::slice(1:3) 

portids <- ports |> dplyr::pull(PORTID)

landings <- gf |> 
  dplyr::filter(Code %in% gfCodes,PORTID %in% portids) |>  
  dplyr::group_by(Year,Area,PORTID) |> 
  dplyr::summarise(lbs = sum(InsideLANDED),
                   .groups = "drop") |>
  tidyr::separate(col=Area,into = c("text","Box"),sep = "ID_") |> 
  dplyr::mutate(Box = as.numeric(Box)) |> 
  dplyr::mutate(mt = lbs*0.000453592) |> 
  dplyr::select(-text,-lbs)

# create map for each
for (iport in portids) {
  
  portName <- ports |> 
    dplyr::filter(PORTID == iport) |> 
    dplyr::pull(newport)
  
  # portlatlon <- ports |> 
  #   dplyr::filter(PORTID == iport)
  
  dataBox <- landings |> 
    dplyr::filter(PORTID == iport) |> 
    dplyr::group_by(Box) |> 
    dplyr::summarise(mt = sum(mt),
                     .groups="drop")
  
  # create landings by box (summed over time)
  boxLandings <- dataBox |>
  dplyr::arrange(desc(mt)) |>
  dplyr::mutate(cumsum = cumsum(mt), prop = cumsum/sum(mt)) |>
  dplyr::filter(prop <= 0.95)

  # select boxes which account fo 95% of landings
  p <- NEFSCspatial::Neus_atlantis |>
    dplyr::left_join(boxLandings,by = c("BOX_ID"="Box"))  |>
    ggplot2::ggplot() +
    ggplot2::geom_sf(ggplot2::aes(fill = mt)) +
    ggplot2::geom_sf_text(ggplot2::aes(label = BOX_ID),
                          size=3) +
    ggplot2::geom_point(data = ports,
                        mapping=ggplot2::aes(x=lon,y=lat),
                        color="red") +
    ggplot2::geom_text(data = ports,
                       ggplot2::aes(x=lon,y=lat,label = newport),
                       size = 2.5,
                       hjust=1, vjust=0) + 
    ggplot2::ggtitle(portName) +
    ggplot2::coord_sf(xlim = c(-80,-62)) 
  
  suppressWarnings(print(p))
}


```

## Location of ports

Location of ports that land gf

```{r ports1, echo =F}
ports <- gf |> 
  dplyr::filter(Code %in% gfCodes) |> 
  dplyr::group_by(newport,PORTID,lat,lon) |> 
  dplyr::summarise(lb = sum(InsideLANDED),
                   .groups = "drop") |> 
  dplyr::arrange(desc(lb)) |> 
  DT::datatable()
ports

```



## Effort

```{r effort1, echo =F}


effort <- gf |> 
  dplyr::filter(Code %in% gfCodes) |> 
  dplyr::group_by(Year,newport) |>
  dplyr::summarise(effort = sum(InsideDAS),
                   lbs = sum(InsideLANDED),
                   .groups = "drop") |> 
  dplyr::filter(newport %in% ports10)
# |> 
#   tidyr::pivot_longer(-c(Year,PORTLANDED),names_to = "type",values_to = "value")

ggplot2::ggplot(effort) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=effort)) +
  ggplot2::facet_wrap(~as.factor(newport)) + 
  ggplot2::ylab("Effort (Days)") +
  ggplot2::ggtitle("Effort by port for top 10 productive ports")

```





