---
title: "NE Groundfish fleet - Effort & Landings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NE Groundfish fleet - Effort & Landings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, warning=FALSE, echo = F, message=F,warning=F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(atlantiscas)
library(ncdf4)
library(sf)
```

## Species

The [NE multispecies (Groundfish) FMP](https://www.nefmc.org/management-plans/northeast-multispecies) comprises the following species:

-   Atlantic Cod (COD)
-   Haddock (HAD)
-   Yellowtail Flounder (YTF)
-   Pollock (POL)
-   American Plaice (PLA)
-   Witch Flounder (WTF)
-   White Hake (WHK)
-   Winter Flounder (WIF)
-   Redfish (RED)
-   Atlantic Halibut (HAL)

Non target species

-   Windowpane flounder (WPF)
-   Ocean Pout (OPT)
-   Atlantic Wolffish (WOL)

These GEARCAT(s) associated with groundfish species are:

-   Bottom Trawl
-   Sink Gillnet

\*\* Still need to separate some species (sharks) \*\*

```{r groundfishdata, echo = F}
# CONVERT TO METRIC TONS
lbstotons <- 2204.62
boundaryBoxes <- c(0,23:29)
threshold1 <- .99 # threshold for filtering data for plots
# gfData <- readRDS(here::here("data/NEGroundfishDataCAMS.rds"))
# gf <- gfData$neus |>
#   dplyr::mutate(Year = as.integer(Year)) |>
#   dplyr::mutate(InsideLANDED = InsideLANDED/lbstotons)

speciesCodes <- c("COD","HAD","YTF","POL","PLA","WTF","WHK","WIF","RED","HAL","WPF","OPT","WOL")

mainPortsIDs = c(240403,240207,220101,240115,420209,240301,320201,350635)
associatedPorts = data.frame(main=c(320201,240115,240301),associated=c(320901,240813,240601))

ports <- unique(c(mainPortsIDs,associatedPorts$associated))


```

## Effort by port

The effort variable is Days at sea (DAS) and is a crude metric. Only integer values and rounded up to the nearest whole day. It is the length of the trip.

After aggregating the top ports based on the decisions found in this section {#sel} and combining all remaining ports into an "other fleet" we can look at the total effort by fleet in units: days at sea

```{r effort1, echo =F}

#this rds file was created using get_effort_landings.r
effortLandByBox <- readRDS(here::here("data/groundfishFleetData.rds")) #get_effort_landings(gf,ports,associatedPorts,speciesCodes)

effortByBox <- effortLandByBox$effort
landingsByBox <- effortLandByBox$landings

annualEffort <- effortByBox |> 
  dplyr::group_by(Year,newport,STATEABB) |> 
  dplyr::summarise(effort = sum(effort),
                   .groups="drop")

ggplot2::ggplot(annualEffort) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=effort)) +
  ggplot2::facet_wrap(~as.factor(newport)) + 
  ggplot2::ylab("Effort (Days)") +
  ggplot2::ggtitle("Effort by port") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```

## Effort by Port

Partition effort by box.

Note: In atlantis effort is entered as as daily effort (days of effort per day). This equates to dividing the annual effort by 365 days.


```{r effortport, echo = F, eval = T}
p <- ggplot2::ggplot(effortByBox) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=effort/365,col = newport)) +
  ggplot2::facet_wrap(~Box) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) +
  ggplot2::ylab("Effort (Days per Day)") +
  ggplot2::guides(col=ggplot2::guide_legend(title="Port"))

plotly::ggplotly(p)


```


# Landings by Port

```{r landingsport, echo = F,eval = T}
#plot landings
boxLandings <- landingsByBox |> 
  dplyr::group_by(Year,Box,newport) |> 
  dplyr::summarise(landings=sum(landings),
                   .groups = "drop")
speciesLandings <- landingsByBox |> 
  dplyr::group_by(Year,Code,newport) |> 
  dplyr::summarise(landings=sum(landings),
                   .groups = "drop")

p <- ggplot2::ggplot(boxLandings) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landings,col = newport)) +
  ggplot2::facet_wrap(~Box) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) +
  ggplot2::ylab("Landings (Metric Tons)") +
  ggplot2::guides(col=ggplot2::guide_legend(title="Port"))

plotly::ggplotly(p)

```


# Landings by Species

```{r landingspecies, echo = F,eval = T}
p <- ggplot2::ggplot(speciesLandings) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landings,col = newport)) +
  ggplot2::facet_wrap(~Code) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) +
  ggplot2::ylab("Landings (Metric Tons)") +
  ggplot2::guides(col=ggplot2::guide_legend(title="Port"))


  plotly::ggplotly(p)
  
```
