---
title: "NE Groundfish fleet - Effort & Landings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NE Groundfish fleet - Effort & Landings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, warning=FALSE, echo = F, message=F,warning=F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(atlantiscas)
library(ncdf4)
library(sf)
source(here::here("R/get_forcing_ts.r"))
```

## Species

The [NE multispecies (Groundfish) FMP](https://www.nefmc.org/management-plans/northeast-multispecies) comprises the following species:

-   Atlantic Cod (COD)
-   Haddock (HAD)
-   Yellowtail Flounder (YTF)
-   Pollock (POL)
-   American Plaice (PLA)
-   Witch Flounder (WTF)
-   White Hake (WHK)
-   Winter Flounder (WIF)
-   Redfish (RED)
-   Atlantic Halibut (HAL)

Non target species

-   Windowpane flounder (WPF)
-   Ocean Pout (OPT)
-   Atlantic Wolffish (WOL)

These GEARCAT(s) associated with groundfish species are:

-   Bottom Trawl
-   Sink Gillnet

\*\* Still need to separate some species (sharks) \*\*

```{r groundfishdata, echo = F}
# CONVERT TO METRIC TONS
lbstotons <- 2204.62
mgNs_to_tonsyr <- .27816 # 1MT/year = .278 mgN/sec
wet_to_C <- 20 # atlantis k_wetdry
C_to_N <- 5.7 # atlantis 

# gfData <- readRDS(here::here("data/NEGroundfishDataCAMS.rds"))
# gf <- gfData$neus |>
#   dplyr::mutate(Year = as.integer(Year)) |>
#   dplyr::mutate(InsideLANDED = InsideLANDED/lbstotons)

speciesCodes <- c("COD","HAD","YTF","POL","PLA","WTF","WHK","WIF","RED","HAL","WPF","OPT","WOL")

mainPortsIDs = c(240403,240207,220101,240115,420209,240301,320201,350635)
associatedPorts = data.frame(main=c(320201,240115,240301),associated=c(320901,240813,240601))

ports <- unique(c(mainPortsIDs,associatedPorts$associated))


```

## Effort by port

The effort variable is Days at sea (DAS) and is a crude metric. Only integer values and rounded up to the nearest whole day. It is the length of the trip.

After aggregating the top ports based on the decisions found in this section {#sel} and combining all remaining ports into an "other fleet" we can look at the total effort by fleet in units: days at sea

```{r effort1, echo =F}

#this rds file was created using get_effort_landings.r
effortLandByBox <- readRDS(here::here("data/groundfishFleetData.rds")) #get_effort_landings(gf,ports,associatedPorts,speciesCodes)

effortByBox <- effortLandByBox$effort
landingsByBox <- effortLandByBox$landings

annualEffort <- effortByBox |> 
  dplyr::group_by(Year,newport,STATEABB) |> 
  dplyr::summarise(effort = sum(effort),
                   .groups="drop")

ggplot2::ggplot(annualEffort) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=effort)) +
  ggplot2::facet_wrap(~as.factor(newport)) + 
  ggplot2::ylab("Effort (Days)") +
  ggplot2::ggtitle("Effort by port") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```

## Effort by Port by box

Partition effort by box.

Note: In atlantis effort is entered as as daily effort (days of effort per day). This equates to dividing the annual effort by 365 days.


```{r effortport, echo = F, eval = T}
p <- ggplot2::ggplot(effortByBox) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=effort/365,col = newport)) +
  ggplot2::facet_wrap(~Box) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) +
  ggplot2::ylab("Effort (Days per Day)") +
  ggplot2::guides(col=ggplot2::guide_legend(title="Port")) +
  ggplot2::theme(panel.spacing.x = ggplot2::unit(1,"lines"),
                 panel.spacing.y = ggplot2::unit(1.5,"lines"))

plotly::ggplotly(p)


```


## Landings by Port

```{r landingsport, echo = F,eval = T}
#plot landings
boxLandings <- landingsByBox |> 
  dplyr::group_by(Year,Box,newport) |> 
  dplyr::summarise(landings=sum(landings),
                   .groups = "drop")
speciesLandings <- landingsByBox |> 
  dplyr::group_by(Year,Code,newport) |> 
  dplyr::summarise(landings=sum(landings),
                   .groups = "drop")

p <- ggplot2::ggplot(boxLandings) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landings,col = newport)) +
  ggplot2::facet_wrap(~Box) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) +
  ggplot2::ylab("Landings (Metric Tons)") +
  ggplot2::guides(col=ggplot2::guide_legend(title="Port")) +
  ggplot2::theme(panel.spacing.x = ggplot2::unit(1,"lines"),
                 panel.spacing.y = ggplot2::unit(1.5,"lines"))

plotly::ggplotly(p)

```


## Landings by Species by port

```{r landingspecies, echo = F,eval = T}
p <- ggplot2::ggplot(speciesLandings) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landings,col = newport)) +
  ggplot2::facet_wrap(~Code) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) +
  ggplot2::ylab("Landings (Metric Tons)") +
  ggplot2::guides(col=ggplot2::guide_legend(title="Port")) +
  ggplot2::theme(panel.spacing.x = ggplot2::unit(1,"lines"),
                 panel.spacing.y = ggplot2::unit(1.5,"lines"))


  plotly::ggplotly(p)
  
```

## Landings by species compare

Compare landings used to force model with landings obtained from CAMS for groundfish fleets.

Note: OPT - ocean pout, WOL - wolfish, WPF - windowpane flounder are not target species. 

Landings in Atlantis are expressed as milligrams of Nitrogen $mgNs^{-1}$. Landings data are in metric Tons of wet weight $mTyr^{-1}$. To convert between the two: 

$$
\begin{aligned}
\text{1 } mgNs^{-1} &= 60*60*24\text{ }mgNday^{-1} \\
&= 365*60*60*24 \text{ }mgN yr^{-1} \\
&= 31536000 * 20 * 5.7 \text{ }mg yr^{-1} \text{(wetweight)} \\
&= 31536000 *20 * 5.7 * 10^{-9}\text{ }MT{yr}^{-1} \\
&= 3.595104 \text{ }MTyr^{-1}
\end{aligned}
$$

where wetweight:C = 20 and C:N = 5.7 (These are set as parameter values in biology.prm as `k_wetdry` and `X_CN` respectively)

```{r landingspeciestot, echo = F,eval = T}
cams <- speciesLandings |> 
  dplyr::group_by(Year,Code) |> 
  dplyr::summarise(landings = sum(landings),
                   .groups = "drop") |> 
  dplyr::mutate(source = "CAMS")

# this is mgN/day
forced <- get_forcing_ts(path="https://raw.githubusercontent.com/NEFSC/READ-EDAB-neusAtlantis/refs/heads/master/currentVersion/CatchFiles/total_catch.ts")

# forced = mgN/second - daily
# Daily input is in mgN/s. Daily value would by mgN/s * 60*60*24
forcedData <- forced |> 
  dplyr::rename(Code = Variable) |> 
  dplyr::filter(Code %in% speciesCodes) |> 
  dplyr::mutate(Year = as.integer(1964+ floor(Time/365))) |> 
  dplyr::group_by(Year,Code) |> 
  # summing mgN/s -> mgN/year
  dplyr::summarise(landings = sum(Value*60*60*24),
                   .groups = "drop") |> 
  # now need to convert mgN/year to mt/year
  # convert N to wet weight (20*5.7) then mg to tons 10^9
  dplyr::mutate(landings = landings*wet_to_C*C_to_N/10^9) |> 
  dplyr::filter(Year >= 1996)  |> 
  dplyr::mutate(source = "Forced")

plotData <- rbind(cams,forcedData) 

p <- plotData |> 
  ggplot2::ggplot() +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landings,color = source)) +
  ggplot2::facet_wrap(~Code,scales = "free_y") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) +
  ggplot2::ylab("Landings (Metric Tons)") +
  ggplot2::guides(col=ggplot2::guide_legend(title="Port")) +
  ggplot2::theme(panel.spacing.x = ggplot2::unit(1,"lines"),
                 panel.spacing.y = ggplot2::unit(1.5,"lines"))

p
```


## Landings of groundfish trips to total

There are some landings of Cod and Haddock that come from other, non groundfish, trips

```{r compare1, echo = F}
# reads in all CAMS data
cleanData <- readRDS(here::here("data-raw/fishing/processedData.rds"))

# split up by groundfish and non groundfish
gftrips <- cleanData$data |>
  dplyr::filter(GEARCAT %in% c("Sink Gillnet","Bottom Trawl"),!squidTrip)

squid <- cleanData$data |>
  dplyr::filter((GEARCAT %in% c("Sink Gillnet","Bottom Trawl") & squidTrip))

other <- cleanData$data |>
  dplyr::filter(!(GEARCAT %in% c("Sink Gillnet","Bottom Trawl")))

## find landings of species in question for each group
gf <- gftrips |>
  dplyr::filter(Code %in% speciesCodes) |>
  dplyr::group_by(Year,Code) |>
  dplyr::summarise(landings = sum(InsideLANDED/lbstotons),
                   .groups = "drop") |>
  dplyr::mutate(fishery = "Groundfish")

ot <- rbind(other,squid) |>
  dplyr::filter(Code %in% speciesCodes) |>
  dplyr::group_by(Year,Code) |>
  dplyr::summarise(landings = sum(InsideLANDED/lbstotons),
                   .groups = "drop") |>
  dplyr::mutate(fishery = "other")

allData <- rbind(gf,ot)

# plot
p <- allData |>
  ggplot2::ggplot()+
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landings,col = fishery))+
  ggplot2::facet_wrap(~Code,scales = "free_y") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) +
  ggplot2::ylab("Landings (Metric Tons)") +
  ggplot2::guides(col=ggplot2::guide_legend(title="Port")) +
  ggplot2::theme(panel.spacing.x = ggplot2::unit(1,"lines"),
                 panel.spacing.y = ggplot2::unit(1.5,"lines"))

p
```
