---
title: "NE Groundfish fleet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NE Groundfish fleet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, warning=FALSE, echo = F, message=F,warning=F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(atlantiscas)
library(ncdf4)
library(sf)
```

## Species

The [NE multispecies (Groundfish) FMP](https://www.nefmc.org/management-plans/northeast-multispecies) comprises the following species:

-   Atlantic Cod (COD)
-   Haddock (HAD)
-   Yellowtail Flounder (YTF)
-   Pollock (POL)
-   American Plaice (PLA)
-   Witch Flounder (WTF)
-   White Hake (WHK)
-   Winter Flounder (WIF)
-   Redfish (RED)
-   Atlantic Halibut (HAL)

Non target species

-   Windowpane flounder (WPF)
-   Ocean Pout (OPT)
-   Atlantic Wolffish (WOL)

These GEARCAT(s) associated with groundfish species are:

-   Bottom Trawl
-   Sink Gillnet

\*\* Still need to separate some species (sharks) \*\*

```{r groundfishdata, echo = F}
# CONVERT TO METRIC TONS
lbstotons <- 2204.62
boundaryBoxes <- c(0,23:29)
threshold1 <- .99 # threshold for filtering data for plots
gfData <- readRDS(here::here("data/NEGroundfishDataCAMS.rds")) 
gf <- gfData$neus |> 
  dplyr::mutate(Year = as.integer(Year)) |> 
  dplyr::mutate(InsideLANDED = InsideLANDED/lbstotons)

speciesCodes <- c("COD","HAD","YTF","POL","PLA","WTF","WHK","WIF","RED","HAL","WPF","OPT","WOL")

gf |> dplyr::distinct(GEARCAT,GEARCODE) |> 
  dplyr::arrange(GEARCAT) |> 
  DT::datatable()

timeRange <- range(gf$Year)
```

## Landings

The landings of trips with associated bycatch

```{r plotland1, echo = F}
landings <- gf |> 
  dplyr::group_by(Year,Functional_Group,GEARCAT) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop")
                     
speciesLanded <- landings |> 
  dplyr::group_by(Functional_Group) |> 
  dplyr::summarise(nyrs = dplyr::n(),
                   .groups = "drop") |> 
  dplyr::filter(nyrs > 1) |> 
  dplyr::pull(Functional_Group)

landings |> 
  dplyr::filter(Functional_Group %in% speciesLanded) |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landed,color=GEARCAT)) +
  ggplot2::facet_wrap(~Functional_Group) +
  ggplot2::ylab("Metric Tons") + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```

## Total groundfish landings for species in FMP

```{r plotcompare, echo = F}
landings <- gf |> 
  dplyr::filter(Code %in% speciesCodes) |> 
  dplyr::mutate(facetLabel = paste0(Functional_Group," (",Code,")")) |> 
  dplyr::group_by(Year,facetLabel,GEARCAT) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop")

p <- landings |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landed,color=GEARCAT)) +
  ggplot2::facet_wrap(~facetLabel)+
  ggplot2::ylab("Metric Tons") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

suppressMessages(print(p))

```

## Total groundfish landings for bycatch species

```{r plotcompare2, echo = F}
landings <- gf |> 
  dplyr::filter(!(Code %in% speciesCodes)) |>  
  dplyr::group_by(Year,Code,GEARCAT) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop")

p <- landings |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landed, color = GEARCAT)) +
  ggplot2::facet_wrap(~Code) +
  ggplot2::ylab("Metric Tons")  +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 
  
suppressMessages(print(p))
  
```

## Landings of groundfish by species

```{r plotland2, echo =F}

gf |> 
  dplyr::group_by(Functional_Group,Code) |> 
  dplyr::summarise(mt = round(sum(InsideLANDED)),
                   .groups = "drop") |> 
  dplyr::arrange(desc(mt)) |> 
  DT::datatable(options = list(pageLength = 10,
                               order = list(list(0, 'asc'))
          ))
```

## Landings of groundfish by box

Spatial maps show Boxes that comprise `r threshold1*100`% of landings for each species. Typically species are caught in all boxes.

```{r landspecies, echo = F, eval = T}

landings <- gf |> 
  dplyr::select(Year,Code,Species,Area,InsideLANDED) |> 
  dplyr::filter(Code %in% speciesCodes) |>  
  tidyr::separate(col=Area,into = c("text","Box"),sep = "ID_") |> 
  dplyr::mutate(Box = as.numeric(Box))

# create map for each
#for (iport in portData$PORTID) {
for (aspecies in speciesCodes) {
  
  boxLandings <- landings |> 
    dplyr::filter(Code == aspecies) |> 
    (\(.) {. ->> sp})() |> 
    dplyr::group_by(Box) |>   
    dplyr::summarise(landed = sum(InsideLANDED),
                     .groups="drop")
    
  Species  <-  sp |> 
    dplyr::pull(Species) |> 
    unique()
  
    # create landings by box (summed over time)
  boxLandings <- boxLandings |>
    dplyr::arrange(desc(landed)) |>
    dplyr::mutate(cumsum = cumsum(landed), prop = cumsum/sum(landed)) |>
    dplyr::filter(prop <= threshold1)

  # select boxes which account fo 95% of landings
  p <- NEFSCspatial::Neus_atlantis |>
    dplyr::left_join(boxLandings,by = c("BOX_ID"="Box"))  |>
    ggplot2::ggplot() +
    ggplot2::geom_sf(ggplot2::aes(fill = landed)) +
    #ggplot2::scale_fill_continuous(limits=c(0, maxBoxLandings))+
    ggplot2::geom_sf_text(ggplot2::aes(label = BOX_ID),
                          size=3) +
    ggplot2::ggtitle(paste0(Species," - (",aspecies,")")) +
    ggplot2::coord_sf(xlim = c(-80,-62)) 
  
  suppressWarnings(print(p))
}


```

## Landings of Groundfish by state

```{r plotland3, echo = F}
landings <- gf |> 
  dplyr::filter(Code %in% speciesCodes) |>  
  dplyr::group_by(Year,STATEABB,GEARCAT) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop")
                     
p <- landings |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landed,color=GEARCAT)) +
  ggplot2::facet_wrap(~STATEABB) +
  ggplot2::ylab("Metric Tons") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

suppressMessages(print(p))

```

## Landings of groundfish by port (species in FMP)

```{r plotland4, echo=F}
landings <- gf |> 
  dplyr::filter(Code %in% speciesCodes) |>  
  dplyr::group_by(Year,STATEABB,newport,GEARCAT) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop")

ports10 <- landings |> 
  dplyr::group_by(newport,STATEABB) |> 
  dplyr::summarise(landed = sum(landed),
                   .groups = "drop") |> 
  dplyr::arrange(desc(landed)) |> 
  dplyr::slice(1:10) |> 
  dplyr::pull(newport)


p <- landings |> 
  dplyr::filter(newport %in% ports10) |> 
  dplyr::group_by(Year,newport,GEARCAT) |> 
  dplyr::summarise(landed = sum(landed),
                   .groups="drop") |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landed, color = GEARCAT)) +
  ggplot2::facet_wrap(~newport) +  
  ggplot2::ylab("Metric Tons") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

suppressMessages(print(p))

```

## Timing of Landings

What is the distribution of effort over the calendar year

```{r gftiming1, echo = F}
gf |> 
  dplyr::select(TRIPID,MONTH,InsideDAS) |> 
  dplyr::distinct() |> 
  dplyr::mutate(MONTH = as.factor(MONTH)) |> 
  dplyr::group_by(MONTH) |> 
  dplyr::summarise(effort = sum(InsideDAS),
                   .groups = "drop") |> 
  ggplot2::ggplot()+
  ggplot2::geom_col(ggplot2::aes(x=MONTH,y=effort)) + 
  ggplot2::ylab("Days at Sea") +
  ggplot2::ggtitle(paste0(range(gf$Year),collapse="-"))



```

## Magnitude of Landings

What is the distribution of landings over the calendar year

```{r gftiming2, echo =F}
gf |> 
  dplyr::filter(Code %in% speciesCodes) |> 
  dplyr::select(MONTH,InsideLANDED) |> 
  dplyr::mutate(MONTH = as.factor(MONTH)) |> 
  dplyr::group_by(MONTH) |> 
  dplyr::summarise(landings = sum(InsideLANDED),
                   .groups = "drop") |> 
  ggplot2::ggplot()+
  ggplot2::geom_col(ggplot2::aes(x=MONTH,y=landings)) + 
  ggplot2::ylab("Metric Tons") +
  ggplot2::ggtitle(paste0(range(gf$Year),collapse="-"))



```

## Landings by Atlantis Box

compared to model output from v2.2.0

```{r plotland5, echo=F}
landings <- gf |> 
  dplyr::filter(Code %in% speciesCodes) |>  
  dplyr::group_by(Year,Area) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop") |>
  tidyr::separate(col=Area,into = c("text","Box"),sep = "ID_") |> 
  dplyr::mutate(Box = as.numeric(Box)) |> 
  dplyr::select(-text) |> 
  dplyr::mutate(source = "CAMS") |> 
  dplyr::mutate(boundary = dplyr::case_when(!(Box %in% boundaryBoxes) ~ "NEUS domain",
                                            .default = "Boundary Box")) 



# Compare to atlantis

# 
# x.nc = nc_open(here::here("data-raw/atlantis",'neus_outputCATCH.nc'))
# varname = names(x.nc$var)
# vn <- grep('SCA_Catch_FC1',varname,value = T)
# 
# dd <- ncvar_get(x.nc,"SCA_Catch_FC1",collapse_degen = F)
# nc_close(x.nc)
# 
# rownames(dd) <- 1:nrow(dd)
# colnames(dd) <- 1:ncol(dd)
# 
# d2 <- as.data.frame(dd) |>
#   tibble::rownames_to_column(var="Box") |>
#   dplyr::mutate(Box = as.integer(Box) -1) |>
#   tidyr::pivot_longer(cols=-Box,names_to = "t",values_to = "value") |>
#   dplyr::mutate(t = as.integer(t),
#                 Year = 1963+ceiling(t/5)) |> 
#   dplyr::group_by(Box,Year) |> 
#   dplyr::summarise(mt = mean(value),
#                    .groups="drop") |> 
#   dplyr::relocate(Year,Box,mt) |> 
#   dplyr::mutate(source = "atlantis")
# 
# dataF <- rbind(d2,landings) |> 
#   dplyr::filter(Year > 1995)
# 

dataF <- landings
dataF |>
  ggplot2::ggplot() +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landed,color=source)) +
  ggplot2::facet_wrap(~Box) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) +
  ggplot2::ylab("Metric Tons") 

```

## Landings (data) by box spatially

Landings are displayed for the Boxes which cumulatively make up 95% of the landings.

```{r plotland6, echo =F}
dataBox <- dataF |> 
  dplyr::filter(source == "CAMS") |> 
  dplyr::group_by(Box) |> 
  dplyr::summarise(landed = sum(landed))

# select boxes which account fo 95% of landings
landings <- dataBox |> 
  dplyr::arrange(desc(landed)) |> 
  dplyr::mutate(cumsum = cumsum(landed), prop = cumsum/sum(landed)) |> 
  dplyr::filter(prop <= 0.95)

p <- NEFSCspatial::Neus_atlantis |> 
  dplyr::left_join(landings,by = c("BOX_ID"="Box"))  |> 
  ggplot2::ggplot() +
  ggplot2::geom_sf(ggplot2::aes(fill = landed)) +
  ggplot2::geom_sf_text(ggplot2::aes(label = BOX_ID),
                        position = ggplot2::position_dodge(width=0.9),  size=3) 

suppressWarnings(print(p))

```

Boxes that contribute 95% of landings `r landings |> dplyr::pull(Box) |> sort()`

## Ports Landings and revenue

Order the ports by NE groundfish landings and revenue.

### Landings

```{r ports1land, echo =F}
ports <- gf |> 
  dplyr::filter(Code %in% speciesCodes) |> 
  dplyr::mutate(lat = round(lat,digits=2),
                lon = round(lon,digits = 2)) |> 
  dplyr::group_by(newport,STATEABB,PORTID,lat,lon) |> 
  dplyr::summarise(mt = round(sum(InsideLANDED)),
                   .groups = "drop")  |> 
  dplyr::arrange(desc(mt)) |> 
  dplyr::mutate(cummt = cumsum(mt),
                percent = round(cummt/sum(mt),digits=2)) |> 
   DT::datatable()
ports

```

### Revenue

Note there are `r gf |> dplyr::filter(is.na(InsideREV)) |> nrow()` records where revenue is NA. Revenue has been deflated to xxxx using the GDP Implicit Price Deflator

```{r ports1rev, echo =F}
portsrev <- gf |> 
  dplyr::filter(Code %in% speciesCodes) |> 
  dplyr::mutate(lat = round(lat,digits=2),
                lon = round(lon,digits = 2)) |> 
  dplyr::group_by(newport,STATEABB,PORTID,lat,lon) |> 
  dplyr::summarise(revenue = round(sum(InsideREV,na.rm = T)),
                   .groups = "drop")  |> 
  dplyr::arrange(desc(revenue)) |> 
  dplyr::mutate(cumrev = cumsum(revenue),
                percent = round(cumrev/sum(revenue),digits=2)) |>
  DT::datatable()
portsrev

```

## Select the main ports {#sel}

The top ports ordered by landings or revenue are in relative agreement. The following ports are selected as the main ports to parameterize in the Atlantis model.

-   MAINE
    -   PORTLAND (220101)
-   NEWHAMPSHIRE
    -   PORTSMOUTH (SEABROOK) (320201, 320901)
-   MASSACHUSETTS
    -   GLOUCESTER (240207)
    -   BOSTON (SCITUATE) (240115, 240813)
    -   CHATHAM (PROVINCETOWN) (240301, 240601)
    -   NEW BEDFORD (240403)
-   RHODE ISLAND
    -   POINT JUDITH (420209)
-   CONNECTICUT
    -   none
-   NEW YORK
    -   MONTAUK (350635)

\*\* Note: MONTAUK, NY added for historic reason (although it was in the top 20 ports)\*\*

```{r mainports, echo = F}
mainPortsIDs = c(240403,240207,220101,240115,420209,240301,320201,350635)
associatedPorts = data.frame(main=c(320201,240115,240301),associated=c(320901,240813,240601))
```

## Landings (data) by box spatially for main ports

Landings are displayed for the Boxes which cumulatively make up `r 100*threshold1`% of the landings.

```{r plotland7, echo =F}

# pick out 3 busiest ports based on landings
ports <- unique(c(mainPortsIDs,associatedPorts$associated))

#  order ports
portids <- NULL
for (iport in mainPortsIDs) {
  portids <- c(portids,iport)
  subports <- associatedPorts |>  
    dplyr::filter(main == iport) |> 
    dplyr::pull(associated)
  portids <- c(portids,subports)    
}


portData <- gf |> 
  dplyr::filter(PORTID %in% ports) |> 
  dplyr::distinct(PORTID,newport,STATEABB,lat,lon) |> 
  dplyr::arrange(desc(lat))


landings <- gf |> 
  dplyr::filter(Code %in% speciesCodes,
                PORTID %in% portData$PORTID) |>  
  dplyr::group_by(Year,Area,PORTID) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop") |>
  tidyr::separate(col=Area,into = c("text","Box"),sep = "ID_") |> 
  dplyr::mutate(Box = as.numeric(Box)) |> 
  dplyr::select(-text)

maxBoxLandings <- landings |> 
  dplyr::group_by(PORTID,Box) |> 
  dplyr::summarise(tots = sum(landed),
                   .groups = "drop") |> 
  dplyr::pull(tots) |> 
  max()

# create map for each
#for (iport in portData$PORTID) {
for (iport in portids) {
  
  portName <- portData |> 
    dplyr::filter(PORTID == iport) |> 
    dplyr::pull(newport)
  
  # portlatlon <- ports |> 
  #   dplyr::filter(PORTID == iport)
  
  dataBox <- landings |> 
    dplyr::filter(PORTID == iport) |> 
    dplyr::group_by(Box) |> 
    dplyr::summarise(landed = sum(landed),
                     .groups="drop")
  
  # create landings by box (summed over time)
  boxLandings <- dataBox |>
    dplyr::arrange(desc(landed)) |>
    dplyr::mutate(cumsum = cumsum(landed), prop = cumsum/sum(landed)) |>
    dplyr::filter(prop <= threshold1)
  
  

  # select boxes which account fo 95% of landings
  p <- NEFSCspatial::Neus_atlantis |>
    dplyr::left_join(boxLandings,by = c("BOX_ID"="Box"))  |>
    ggplot2::ggplot() +
    ggplot2::geom_sf(ggplot2::aes(fill = landed)) +
    ggplot2::scale_fill_continuous(limits=c(0, maxBoxLandings))+
    ggplot2::geom_sf_text(ggplot2::aes(label = BOX_ID),
                          size=3) +
    ggplot2::geom_point(data = portData,
                        mapping=ggplot2::aes(x=lon,y=lat),
                        color="green") +
    ggplot2::geom_point(data = portData |> dplyr::filter(PORTID == iport),
                        mapping=ggplot2::aes(x=lon,y=lat),
                        color="red") +

    ggplot2::geom_text(data = portData,
                       ggplot2::aes(x=lon,y=lat,label = newport),
                       size = 2.5,
                       hjust=1, vjust=0) + 
    ggplot2::ggtitle(portName) +
    ggplot2::coord_sf(xlim = c(-80,-62)) 
  
  suppressWarnings(print(p))
}


```

## Aggregate remaining data as "other" fleet

The remaining ports (other than the ports listed above) make up \~10% of total landings. All trips associated with these ports will be combined to form one fleet which will fish in the footprint below. This maintains the communities at sea idea for the main ports but takes into account all remaining landings to satisfy model requirements

```{r otherfleet,echo = F}
ports <- unique(c(mainPortsIDs,associatedPorts$associated))

otherFleet <- gf |> 
  dplyr::filter(!(PORTID %in% ports),
                Code %in% speciesCodes) |> 
  dplyr::group_by(Year,Area) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop") |>
  tidyr::separate(col=Area,into = c("text","Box"),sep = "ID_") |> 
  dplyr::mutate(Box = as.numeric(Box)) |> 
  dplyr::select(-text)

otherFleetLandings <- sum(otherFleet$landed)
timerange <- range(otherFleet$Year)

otherFleetBox <- otherFleet |> 
  dplyr::group_by(Box) |> 
  dplyr::summarise(landed = sum(landed),
                   .groups = "drop") 

  
  # select boxes which account fo 95% of landings
  p <- NEFSCspatial::Neus_atlantis |>
    dplyr::left_join(otherFleetBox,by = c("BOX_ID"="Box"))  |>
    ggplot2::ggplot() +
    ggplot2::geom_sf(ggplot2::aes(fill = landed)) +
    #ggplot2::scale_fill_continuous(limits=c(0, maxBoxLandings))+
    ggplot2::geom_sf_text(ggplot2::aes(label = BOX_ID),
                          size=3) +
    ggplot2::ggtitle("Other Fleet") +
    ggplot2::coord_sf(xlim = c(-80,-62)) 
  
  suppressWarnings(print(p))

```

Total Landings over `r paste(timerange,collapse="-")` = `r round(otherFleetLandings)` metric tons

Time series of landings by box of other fleet

```{r otherfleettime, echo = F}

otherFleetBoundary <- 
  otherFleet |> 
  dplyr::mutate(boundary = dplyr::case_when(!(Box %in% boundaryBoxes) ~ "NEUS domain",
                                            .default = "Boundary Box")) 
  
otherFleetBoundary |> 
  ggplot2::ggplot() +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landed,color = factor(boundary)))+
  ggplot2::facet_wrap(~Box) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) +
  ggplot2::ylab("Metric Tons") +
  ggplot2::labs(color=" ")
```

## Effort

The effort variable is Days at sea (DAS) and is a crude metric. Only integer values and rounded up to the nearest whole day. It is the length of the trip.

After aggregating the top ports based on the decisions found in this section {#sel} and combining all remaining ports into an "other fleet" we can look at the total effort by fleet in units: days at sea

```{r effort1, echo =F}

ports <- unique(c(mainPortsIDs,associatedPorts$associated))

effortLandByBox <- get_effort_landings(gf,ports,associatedPorts,speciesCodes)

effortByBox <- effortLandByBox$effort
landingsByBox <- effortLandByBox$landings

annualEffort <- effortByBox |> 
  dplyr::group_by(Year,newport,STATEABB) |> 
  dplyr::summarise(effort = sum(effort),
                   .groups="drop")

ggplot2::ggplot(annualEffort) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=effort)) +
  ggplot2::facet_wrap(~as.factor(newport)) + 
  ggplot2::ylab("Effort (Days)") +
  ggplot2::ggtitle("Effort by port") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```

## Effort by box (Time series)

Partition effort by box.

Note: In atlantis effort is entered as as daily effort (days of effort per day). This equates to dividing the annual effort by 365 days. Her we show the effort by box for the top 3 ports (by landings)

```{r effort2, echo =F}

port3 <- head(ports10,3)

boxEffort <- effortByBox |> 
  dplyr::filter(newport %in% port3)

for (aport in port3) {
  p <- ggplot2::ggplot(boxEffort |> dplyr::filter(newport == aport)) +
    ggplot2::geom_line(ggplot2::aes(x=Year,y=effort)) +
    ggplot2::facet_wrap(~as.factor(Box)) + 
    ggplot2::ylab("Effort (Days)") +
    ggplot2::ggtitle(aport) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 
  
  suppressMessages(print(p))
}

```

## Landings by box (Time series)

```{r gflandingsboxtime, echo =F}

landBox<- landingsByBox |> 
  dplyr::filter(newport %in% port3) |> 
  dplyr::group_by(Year,Box,newport) |> 
  dplyr::summarise(landings = sum(landings),
                   .groups="drop")

for (aport in port3) {
  p <- ggplot2::ggplot(landBox |> dplyr::filter(newport == aport)) +
    ggplot2::geom_line(ggplot2::aes(x=Year,y=landings)) +
    ggplot2::facet_wrap(~as.factor(Box)) + 
    ggplot2::ylab("Metric Tons") +
    ggplot2::ggtitle(aport) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 
  
 suppressMessages(print(p))

}

```
