---
title: "NE Groundfish selectivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NE Groundfish selectivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, warning=FALSE, echo = F, message=F,warning=F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(atlantiscas)
library(ncdf4)
library(sf)
```

## Species

The [NE multispecies (Groundfish) FMP](https://www.nefmc.org/management-plans/northeast-multispecies) 

-   Atlantic Cod (COD)
-   Haddock (HAD)
-   Yellowtail Flounder (YTF)
-   Pollock (POL)
-   American Plaice (PLA)
-   Witch Flounder (WTF)
-   White Hake (WHK)
-   Winter Flounder (WIF)
-   Redfish (RED)
-   Atlantic Halibut (HAL)

Non target species

-   Windowpane flounder (WPF)
-   Ocean Pout (OPT)
-   Atlantic Wolffish (WOL)

```{r groundselectivity, echo = F,eval=T, message=F}
# pull groundfish data
lbstotons <- 2204.62
boundaryBoxes <- c(0,23:29)
# species of interest
gfcodes <- c("RED", "WHK", "COD" ,"HAD" ,"POL", "WTF", "WIF", "HAL", "PLA", "YTF" ,"WPF" ,"WOL", "OPT")

gfData <- readRDS(here::here("data/NEGroundfishDataCAMS.rds"))
gf <- gfData$neus |>
  dplyr::mutate(Year = as.integer(Year)) |>
  dplyr::mutate(InsideLANDED = InsideLANDED/lbstotons)

# reads in CAMS tables
if(0) {
  # attempts to join tables with GARFO
  channel <- dbutils::connect_to_database("NEFSC_USERS","abeet")

  sql1 <- "select distinct CAMSID, VTR_IMGID
  from CAMS_GARFO.CAMS_SUBTRIP
  where YEAR >= 1996
  and VTR_IMGID IS NOT NULL"
  subtrip <- DBI::dbGetQuery(channel,sql1)

  # this table is a list of camsid and original vtr ids
  saveRDS(subtrip,here::here("data-raw/subtrip.rds"))
}

# distinct camsid, vtr_imgid pairs
subtrip <- readRDS(here::here("data-raw/subtrip.rds")) |>
  dplyr::as_tibble() |>
  dplyr::mutate(VTR_IMGID = as.character(VTR_IMGID))

# select trips
# IDNUM in our data = VTR_IMGID
trips <- gf |>
  dplyr::distinct(IDNUM) |>
  dplyr::pull()

# trips in groundfish data
pairedtrips <-  subtrip |>
  dplyr::filter(VTR_IMGID %in% trips)

# sel;ect trip id and gear type
gftripgear <- gf |>
  dplyr::select(IDNUM,GEARCAT) |>
  dplyr::distinct()

# find associated cams ids so we can get lengths by gear cat
idsbyGear <- gftripgear |>
  dplyr::left_join(pairedtrips,by=c("IDNUM"="VTR_IMGID")) |>
  dplyr::filter(!is.na(CAMSID))

camstrips <- idsbyGear |>
  dplyr::pull(CAMSID)


# now pull lengths of species caught 
if(0) {
  sqlLengths <- "select CAMSID, NESPP3, LENGTH, NUMLEN
  from CAMS_GARFO.cams_cflen_aa
  where YEAR >= 1996"

  lengths <- DBI::dbGetQuery(channel,sqlLengths) |>
    dplyr::as_tibble()
}

lengths <- readRDS(here::here("data-raw/gflengths.rds"))

# filter trips based on GF trips
gflengths <- lengths |>
  dplyr::filter(CAMSID %in% camstrips) |>
  dplyr::mutate(NESPP3 = as.double(NESPP3))

# reads in atlantis groups file to find nespp3 codes
atlantisGroups <- readr::read_csv("https://raw.githubusercontent.com/NEFSC/READ-EDAB-neusAtlantis/dev_branch/data-raw/data/Atlantis_2_0_groups_svspp_nespp3.csv",show_col_types = FALSE)
gfgroups <- atlantisGroups |>
  dplyr::filter(Code %in% gfcodes)

## join by nespp3
allgflengths <- gflengths |>
  dplyr::left_join(gfgroups,by = "NESPP3") |>
  dplyr::filter(Code %in% gfcodes) |> 
  dplyr::filter(LENGTH < 200)

# replicate rows based on NUMLEN field
# eg LENGTH = 46 , NUMLE = 6. 6 fish caught of length 46
# AMERICAN  PLAICE have anomolous large lengths. these are removed
replicatelengths <- allgflengths |>
  tidyr::uncount(NUMLEN) 
```

## Histogram of lengths

Number of fish over all designated groundfish fleets

```{r sel1,echo = F, eval = T}

# Plot over all fleets.
# to do this by fleet we 'd need to filter out tripid's specific to designated fleets
# plot histogram of lengths
replicatelengths |>
  ggplot2::ggplot() +
  ggplot2::geom_histogram(ggplot2::aes(x=LENGTH),bins=25) +
  ggplot2::facet_wrap(~Code,scale = "free_x") +
  ggplot2::ylab("Number of fish") +
  ggplot2::xlab("length (cm)")
```

## Probability curves

Fish over all designated groundfish fleets

```{r sel2, echo=F, eval=T}
# plot prob curve by species
replicatelengths |>
  dplyr::select(Code,LENGTH) |>
  dplyr::arrange(Code,LENGTH) |>
  dplyr::group_by(Code) |>
  dplyr::mutate(prob = (1:dplyr::n())/dplyr::n()) |>
  ggplot2::ggplot(ggplot2::aes(x=LENGTH,y=prob)) +
  ggplot2::geom_line() +
  ggplot2::geom_point(size = 0.5) +
  ggplot2::facet_wrap(~Code,scale = "free_x") +
  ggplot2::ylab("Probability") +
  ggplot2::xlab("length (cm)")
```

## Probability curves stacked

```{r sel3 ,echo=F,eval=T}
# plot all on same figure
p <- replicatelengths |>
  dplyr::select(Code,LENGTH) |>
  dplyr::arrange(Code,LENGTH) |>
  dplyr::group_by(Code) |>
  dplyr::mutate(prob = (1:dplyr::n())/dplyr::n()) |>
  ggplot2::ggplot(ggplot2::aes(x=LENGTH,y=prob, col=Code)) +
  ggplot2::geom_line() +
  ggplot2::geom_point(size = 0.5) +
  ggplot2::ylab("Probability") +
  ggplot2::xlab("length (cm)")

plotly::ggplotly(p)

```

## Selectivities by GEARCAT

```{r selbygear,echo=F}
## separate by gear type and repeat plots to
# determine if there is a difference in selectivity based on
# fishing gear
gill <- idsbyGear |>
  dplyr::filter(GEARCAT == "Sink Gillnet") |>
  dplyr::distinct(GEARCAT,CAMSID)
trawl <- idsbyGear |>
  dplyr::filter(GEARCAT == "Bottom Trawl") |>
  dplyr::distinct(GEARCAT,CAMSID)

# NAs introduced from CAMS trips associated with other GEARCAT type
gilllengths <- allgflengths |>
  dplyr::left_join(gill,by = "CAMSID") |>
  tidyr::uncount(NUMLEN) |> 
  dplyr::filter(!is.na(GEARCAT))
trawllengths <- allgflengths |>
  dplyr::left_join(trawl,by = "CAMSID") |>
  tidyr::uncount(NUMLEN) |> 
  dplyr::filter(!is.na(GEARCAT))

combinedlengths <- rbind(gilllengths,trawllengths) |>
  dplyr::filter(!is.na(GEARCAT))
```

### Bottom Trawl

```{r plotTrawl, echo = F}
# plot selectivities by gear
trawllengths |>
  dplyr::select(Code,LENGTH) |>
  dplyr::arrange(Code,LENGTH) |>
  dplyr::group_by(Code) |>
  dplyr::mutate(prob = (1:dplyr::n())/dplyr::n()) |>
  ggplot2::ggplot(ggplot2::aes(x=LENGTH,y=prob)) +
  ggplot2::geom_line() +
  ggplot2::geom_point(size = 0.5) +
  ggplot2::facet_wrap(~Code,scale = "free_x") +
  ggplot2::ylab("Probability") +
  ggplot2::xlab("length (cm)")
```

### Gill net

```{r plotGill, echo = F}
gilllengths |>
  dplyr::select(Code,LENGTH) |>
  dplyr::arrange(Code,LENGTH) |>
  dplyr::group_by(Code) |>
  dplyr::mutate(prob = (1:dplyr::n())/dplyr::n()) |>
  ggplot2::ggplot(ggplot2::aes(x=LENGTH,y=prob)) +
  ggplot2::geom_line() +
  ggplot2::geom_point(size = 0.5) +
  ggplot2::facet_wrap(~Code,scale = "free_x") +
  ggplot2::ylab("Probability") +
  ggplot2::xlab("length (cm)")
```


### Combined

```{r plotCombibned, echo = F}
# plot prob curve by species
combinedlengths |>
  dplyr::select(Code,GEARCAT,LENGTH) |>
  dplyr::arrange(Code,GEARCAT,LENGTH) |>
  dplyr::group_by(Code,GEARCAT) |>
  dplyr::mutate(prob = (1:dplyr::n())/dplyr::n()) |>
  ggplot2::ggplot(ggplot2::aes(x=LENGTH,y=prob,col=GEARCAT)) +
  ggplot2::geom_line() +
  ggplot2::geom_point(size = 0.5) +
  ggplot2::facet_wrap(~Code,scale = "free_x") +
  ggplot2::ylab("Probability") +
  ggplot2::xlab("length (cm)")

```
