---
title: "scallops"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scallops}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, echo = F, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(atlantiscas)
library(ncdf4)
library(sf)
```

## Atlantis parameterization
(Codebase v6681)

1. Forced effort time series (days $day^{-1}$) - effortmodel = 11 (Running into segmentation error issues)

2. YYY_effort > 0 (Fixed total effort, days $day^{-1}$)


## Trial for Atlantis conversion

First look at the data to see how many gear types, and which ports landed the scallops.

The fishing data were processed and passed through minimal QA/QC. All ports were assigned lat and lons (as best as possible), all landings were assigned to an atlantis group.

SCA is the atlantis group for scallops.

Gear Categories identified, in the data set, as scallop vessels were

* TRAWL BOTTOM SCALLOP
* DREDGE SCALLOP
* Scallop Gear 

These GEARCAT(s) were filtered to provide the following output

```{r scallopData, echo = F}
lbstotons <- 2204.62
boundaryBoxes <- c(0,23:29)
scallopData <- readRDS(here::here("data/scallopDataCAMS.rds"))

sca <- scallopData$neus |> 
  dplyr::mutate(Year = as.integer(Year),
                InsideLANDED = InsideLANDED/lbstotons)

sca |> dplyr::distinct(GEARCAT,GEARCODE) |> 
  dplyr::arrange(GEARCAT) |> 
  DT::datatable()
```

## Landings (All species)

The landings of Scallop trips with associated bycatch

```{r plotland1, echo = F}
landings <- sca |> 
  dplyr::group_by(Year,Functional_Group) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop")
                     
speciesLanded <- landings |> 
  dplyr::group_by(Functional_Group) |> 
  dplyr::summarise(nyrs = dplyr::n(),
                   .groups = "drop") |> 
  dplyr::filter(nyrs > 1) |> 
  dplyr::pull(Functional_Group)

p <- landings |> 
  dplyr::filter(Functional_Group %in% speciesLanded) |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landed)) +
  ggplot2::facet_wrap(~Functional_Group) +
  ggplot2::ylab("Metric Tons") + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

suppressMessages(print(p))


```

## Total scallop landings compared to bycatch

```{r plotland2, echo =F}

sca |> 
  dplyr::group_by(Functional_Group) |> 
  dplyr::summarise(landed = round(sum(InsideLANDED),digits=2),
                   .groups = "drop") |> 
  dplyr::arrange(desc(landed)) |> 
  DT::datatable(options = list(pageLength = 10,
                               order = list(list(0, 'asc'))
          ))
```


## Landings of Scallops by state

```{r plotland3, echo = F}
landings <- sca |> 
  dplyr::filter(Code == "SCA") |>  
  dplyr::group_by(Year,STATEABB) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop")
                     
landings |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landed)) +
  ggplot2::facet_wrap(~STATEABB) +
  ggplot2::ylab("Metric Tons") + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90))  

```

## Landings of Scallops by port

```{r plotland4, echo=F}
landings <- sca |> 
  dplyr::filter(Code == "SCA") |>  
  dplyr::group_by(Year,STATEABB,newport) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop")

ports10 <- landings |> 
  dplyr::group_by(newport,STATEABB) |> 
  dplyr::summarise(landed = sum(landed),
                   .groups = "drop") |> 
  dplyr::arrange(desc(landed)) |> 
  dplyr::slice(1:10) |> 
  dplyr::pull(newport)


landings |> 
  dplyr::filter(newport %in% ports10) |> 
  dplyr::group_by(Year,newport) |> 
  dplyr::summarise(landed = sum(landed),
                   .groups="drop") |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landed)) +
  ggplot2::facet_wrap(~newport) +
  ggplot2::ylab("Metric Tons") + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```

```{r portland4b, echo=F}
ports <- sca |> 
  dplyr::group_by(newport,STATEABB,PORTID,lat,lon) |> 
  dplyr::summarise(mt = sum(InsideLANDED),
                   .groups = "drop") |> 
  dplyr::mutate(lat = round(lat,digits=2),
                lon = round(lon,digits = 2)) |> 
  dplyr::arrange(desc(mt)) |> 
  dplyr::mutate(cummt = cumsum(mt),
                percent = round(cummt/sum(mt),digits=2)) 

DT::datatable(ports)
```


## Select the main ports {#sel}

The following ports three main ports are selected. Other smaller ports (listed under main port) are combined with main ports. hese ports comprise ~ 90% of total landings. 

* MASSACHUSETTS
  * NEW BEDFORD (240403)
    * FAIRHAVEN (242203)
* NEW JERSEY
  * CAPE MAY (330309)
    * LONG BEACH (331627)
    * POINT PLEASANT (330127)
* VIRGINIA
  * NEWPORT NEWS (490910)
    * SEAFORD (490869)
    * HAMPTON (490118)
    
```{r mainports, echo = F}
mainPortsIDs = c(240403,330309,490910)
associatedPorts = data.frame(main=c(240403,330309,330309,490910,490910),associated=c(242203,331627,330127,490869,490118))
threshold1 <- .99
```


## Landings by Atlantis Box
compared to model output from v2.2.0

```{r plotland5, echo=F}
landings <- sca |> 
  dplyr::filter(Code == "SCA") |>  
  dplyr::group_by(Year,Area) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop") |>
  tidyr::separate(col=Area,into = c("text","Box"),sep = "ID_") |> 
  dplyr::mutate(Box = as.numeric(Box)) |> 
  dplyr::select(-text) |> 
  dplyr::mutate(source = "CAMS")

# Atlantis v2.2.0 scallop catch by box

x.nc = nc_open(here::here("data-raw/atlantis",'neus_outputCATCH.nc'))
varname = names(x.nc$var)
vn <- grep('SCA_Catch_FC1',varname,value = T)

dd <- ncvar_get(x.nc,"SCA_Catch_FC1",collapse_degen = F)
nc_close(x.nc)

rownames(dd) <- 1:nrow(dd)
colnames(dd) <- 1:ncol(dd)

d2 <- as.data.frame(dd) |>
  tibble::rownames_to_column(var="Box") |>
  dplyr::mutate(Box = as.integer(Box) -1) |>
  tidyr::pivot_longer(cols=-Box,names_to = "t",values_to = "value") |>
  dplyr::mutate(t = as.integer(t),
                Year = 1963+ceiling(t/5)) |> 
  dplyr::group_by(Box,Year) |> 
  dplyr::summarise(landed = mean(value),
                   .groups="drop") |> 
  dplyr::relocate(Year,Box,landed) |> 
  dplyr::mutate(source = "atlantis")

dataF <- rbind(d2,landings) |> 
  dplyr::filter(Year > 1995)
# 
# landings |> 
#   ggplot2::ggplot() + 
#   ggplot2::geom_line(ggplot2::aes(x=Year,y=mt)) +
#   ggplot2::facet_wrap(~Box) +
#   ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90))

dataF |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landed,color=source)) +
  ggplot2::facet_wrap(~Box) +
  ggplot2::ylab("Metric Tons") + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```

## Landings (data) by box spatially

```{r plotland6, echo =F}
dataBox <- dataF |> 
  dplyr::filter(source == "CAMS") |> 
  dplyr::group_by(Box) |> 
  dplyr::summarise(landed = sum(landed))

# select boxes which account fo 95% of landings
landings <- dataBox |> 
  dplyr::arrange(desc(landed)) |> 
  dplyr::mutate(cumsum = cumsum(landed), prop = cumsum/sum(landed)) |> 
  dplyr::filter(prop <= 0.95)

NEFSCspatial::Neus_atlantis |> 
  dplyr::left_join(landings,by = c("BOX_ID"="Box"))  |> 
  ggplot2::ggplot() +
  ggplot2::geom_sf(ggplot2::aes(fill = landed)) +
  ggplot2::geom_sf_text(ggplot2::aes(label = BOX_ID),
                        position = ggplot2::position_dodge(width=0.9),  size=3)  +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```

Boxes that contribute 95% of landings `r landings |> dplyr::pull(Box) |> sort()`


## Landings (data) by box spatially for main ports

```{r plotland7, echo =F}

	
portids <- unique(c(mainPortsIDs,associatedPorts$associated))

ports <- sca |> 
  dplyr::group_by(newport,STATEABB,PORTID,lat,lon) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop") |> 
  dplyr::arrange(desc(landed)) |> 
  dplyr::filter(PORTID %in% portids) 

landings <- sca |> 
  dplyr::filter(Code == "SCA",PORTID %in% portids) |>  
  dplyr::group_by(Year,Area,PORTID) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop") |>
  tidyr::separate(col=Area,into = c("text","Box"),sep = "ID_") |> 
  dplyr::mutate(Box = as.numeric(Box)) |> 
  dplyr::select(-text)


maxBoxLandings <- landings |> 
  dplyr::group_by(PORTID,Box) |> 
  dplyr::summarise(tots = sum(landed),
                   .groups = "drop") |> 
  dplyr::pull(tots) |> 
  max()

# create map for each
for (iport in portids) {
  
  portName <- ports |> 
    dplyr::filter(PORTID == iport) |> 
    dplyr::pull(newport)
  
  # portlatlon <- ports |> 
  #   dplyr::filter(PORTID == iport)
  
  dataBox <- landings |> 
    dplyr::filter(PORTID == iport) |> 
    dplyr::group_by(Box) |> 
    dplyr::summarise(landed = sum(landed),
                     .groups="drop")
  
  # create landings by box (summed over time)
  boxLandings <- dataBox |>
  dplyr::arrange(desc(landed)) |>
  dplyr::mutate(cumsum = cumsum(landed), prop = cumsum/sum(landed)) |>
  dplyr::filter(prop <= 0.95)

  # select boxes which account fo 95% of landings
  p <- NEFSCspatial::Neus_atlantis |>
    dplyr::left_join(boxLandings,by = c("BOX_ID"="Box"))  |>
    ggplot2::ggplot() +
    ggplot2::geom_sf(ggplot2::aes(fill = landed)) +
    ggplot2::scale_fill_continuous(limits=c(0, maxBoxLandings))+
    ggplot2::geom_sf_text(ggplot2::aes(label = BOX_ID),
                          size=3) +
    ggplot2::geom_point(data = ports,
                        mapping=ggplot2::aes(x=lon,y=lat),
                        color="green") +
    ggplot2::geom_point(data = ports |> dplyr::filter(PORTID == iport),
                        mapping=ggplot2::aes(x=lon,y=lat),
                        color="red") +
    ggplot2::geom_text(data = ports,
                       ggplot2::aes(x=lon,y=lat,label = newport),
                       size = 2.5,
                       hjust=1, vjust=0) + 
    ggplot2::ggtitle(portName) +
    ggplot2::coord_sf(xlim = c(-80,-62)) 
  
  suppressWarnings(print(p))
}


```

## Aggregate remaining data as "other" fleet

The remaining ports (other than the ports listed above) make up ~10% of total landings. All trips associated with these ports will be combined to form one fleet which will fish in the footprint below. This maintains the communities at sea idea for the main ports but takes into account all remaining landings to satisfy model requirements

```{r otherfleet,echo = F}
ports <- unique(c(mainPortsIDs,associatedPorts$associated))

otherFleet <- sca |> 
  dplyr::filter(!(PORTID %in% ports)) |> 
  dplyr::group_by(Year,Area) |> 
  dplyr::summarise(landed = sum(InsideLANDED),
                   .groups = "drop") |>
  tidyr::separate(col=Area,into = c("text","Box"),sep = "ID_") |> 
  dplyr::mutate(Box = as.numeric(Box)) |> 
  dplyr::select(-text)

otherFleetLandings <- sum(otherFleet$landed)
timerange <- range(otherFleet$Year)

otherFleetBox <- otherFleet |> 
  dplyr::group_by(Box) |> 
  dplyr::summarise(landed = sum(landed),
                   .groups = "drop") 

  
  # select boxes which account fo 95% of landings
  p <- NEFSCspatial::Neus_atlantis |>
    dplyr::left_join(otherFleetBox,by = c("BOX_ID"="Box"))  |>
    ggplot2::ggplot() +
    ggplot2::geom_sf(ggplot2::aes(fill = landed)) +
    #ggplot2::scale_fill_continuous(limits=c(0, maxBoxLandings))+
    ggplot2::geom_sf_text(ggplot2::aes(label = BOX_ID),
                          size=3) +
    ggplot2::ggtitle("Other Fleet") +
    ggplot2::coord_sf(xlim = c(-80,-62)) 
  
  suppressWarnings(print(p))

```

Total Landings over `r paste(timerange,collapse="-")` = `r round(otherFleetLandings)` metric tons

Time series of landings by box of other fleet


```{r otherfleettime, echo = F, message=F}

otherFleetBoundary <- 
  otherFleet |> 
  dplyr::mutate(boundary = dplyr::case_when(!(Box %in% boundaryBoxes) ~ "NEUS domain",
                                            .default = "Boundary Box")) 
  
p <- otherFleetBoundary |> 
  ggplot2::ggplot() +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=landed,color = factor(boundary)))+
  ggplot2::facet_wrap(~Box) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) +
  ggplot2::ylab("Metric Tons") +
  ggplot2::labs(color=" ")

suppressMessages(print(p))
```


## Effort

The effort variable is Days at sea (DAS) and is a crude metric. Only integer values and rounded up to the nearest whole day. It is the length of the trip. Note: For each record of a trip (multiple records based on the species caught) the DAS is replicated, so to calculate the true DAS we need to find distinct Trip-DAS

After aggregating the top ports based on the decisions found in this section {#sel} and combining all remaining ports into an "other fleet" we can look at the total effort by fleet in units: days at sea


```{r effort1, echo =F}

ports <- unique(c(mainPortsIDs,associatedPorts$associated))

effortByBox <- get_effort(sca,ports,associatedPorts)

annualEffort <- effortByBox |> 
  dplyr::group_by(Year,newport,STATEABB) |> 
  dplyr::summarise(effort = sum(effort),
                   .groups="drop")

ggplot2::ggplot(annualEffort) +
  ggplot2::geom_line(ggplot2::aes(x=Year,y=effort)) +
  ggplot2::facet_wrap(~as.factor(newport)) + 
  ggplot2::ylab("Effort (Days)") +
  ggplot2::ggtitle("Effort by port") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 

```

## Effort by box (Time series)

Partition effort by box.

Note: In atlantis effort is entered as as daily effort (days of effort per day). This equates to dividing the annual effort by 365 days. Here we show the effort by box for the top ports (by landings)

```{r effort2, echo =F}

port3 <-effortByBox |> 
  dplyr::distinct(newport) |> 
  dplyr::pull()

boxEffort <- effortByBox |> 
  dplyr::filter(newport %in% port3)

for (aport in port3) {
  p <- ggplot2::ggplot(boxEffort |> dplyr::filter(newport == aport)) +
    ggplot2::geom_line(ggplot2::aes(x=Year,y=effort)) +
    ggplot2::facet_wrap(~as.factor(Box)) + 
    ggplot2::ylab("Effort (Days)") +
    ggplot2::ggtitle(aport) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 
  
  suppressMessages(print(p))

}

```

## Landings by box (Time series)

```{r landingsboxtime, echo =F}

boxEffort <- effortByBox |> 
  dplyr::filter(newport %in% port3)

for (aport in port3) {
  p <- ggplot2::ggplot(boxEffort |> dplyr::filter(newport == aport)) +
    ggplot2::geom_line(ggplot2::aes(x=Year,y=landings)) +
    ggplot2::facet_wrap(~as.factor(Box)) + 
    ggplot2::ylab("Metric Tons") +
    ggplot2::ggtitle(aport) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90)) 
  
 suppressMessages(print(p))

}

```


