---
title: "scallops"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scallops}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(atlantiscas)
```

## Trial for Atlantis conversion

First look at the data to see how many gear types, and which ports landed the scallops.

The fishing data were processed and passed through minimal QA/QC. All ports were assigned lat and lons (as best as possible), all landings were assigned to an atlantis group.

SCA is the atlantis group for scallops.

Gear Categories identified, in the data set, as scallop vessels were

* TRAWL BOTTOM SCALLOP
* DREDGE SCALLOP
* Scallop Gear 

These GEARCAT(s) were filtered to provide the following output

```{r scallopData, echo = F}
sca <- readRDS(here::here("data/scallopData.rds")) |> 
  dplyr::mutate(Year = as.integer(Year))

sca |> dplyr::distinct(GEARCAT,GEARCODE) |> 
  dplyr::arrange(GEARCAT) |> 
  knitr::kable()
```

## Landings

The landings of Scallops and associated bycatch

```{r plotland1, echo = F}
landings <- sca |> 
  dplyr::group_by(Year,Functional_Group) |> 
  dplyr::summarise(lbs = sum(InsideLANDED),
                   .groups = "drop")
                     
speciesLanded <- landings |> 
  dplyr::group_by(Functional_Group) |> 
  dplyr::summarise(nyrs = dplyr::n(),
                   .groups = "drop") |> 
  dplyr::filter(nyrs > 1) |> 
  dplyr::pull(Functional_Group)

landings |> 
  dplyr::filter(Functional_Group %in% speciesLanded) |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=lbs)) +
  ggplot2::facet_wrap(~Functional_Group)

```

Total scallop landings compared with atlantis forcing time series

```{r plotcompare, echo = F}
landings <- sca |> 
  dplyr::filter(Code == "SCA") |>  
  dplyr::group_by(Year,STATE) |> 
  dplyr::summarise(lbs = sum(InsideLANDED),
                   .groups = "drop")

```


Table of landings

```{r plot land2, echo =F}

sca |> 
  dplyr::group_by(Functional_Group) |> 
  dplyr::summarise(lbs = sum(InsideLANDED),
                   .groups = "drop") |> 
  dplyr::arrange(desc(lbs)) |> 
  kableExtra::kable()
```


Landings by state

```{r plotland3, echo = F}
landings <- sca |> 
  dplyr::filter(Code == "SCA") |>  
  dplyr::group_by(Year,STATE) |> 
  dplyr::summarise(lbs = sum(InsideLANDED),
                   .groups = "drop")
                     
landings |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=lbs)) +
  ggplot2::facet_wrap(~STATE)

```

Landings by port

```{r plotland4, echo=F}
landings <- sca |> 
  dplyr::filter(Code == "SCA") |>  
  dplyr::group_by(Year,STATE,PORTLANDED) |> 
  dplyr::summarise(lbs = sum(InsideLANDED),
                   .groups = "drop")

ports <- landings |> 
  dplyr::group_by(PORTLANDED,STATE) |> 
  dplyr::summarise(lbs = sum(lbs),
                   .groups = "drop") |> 
  dplyr::arrange(desc(lbs)) |> 
  dplyr::slice(1:10) |> 
  dplyr::pull(PORTLANDED)


landings |> 
  dplyr::filter(PORTLANDED %in% ports) |> 
  ggplot2::ggplot() + 
  ggplot2::geom_line(ggplot2::aes(x=Year,y=lbs)) +
  ggplot2::facet_wrap(~PORTLANDED)

```


